---
title: "Lecture 4: Clustering & Unsupervised Learning"
subtitle: "Pattern Discovery Without Labels"
author: "Predictive Analytics & Big Data - Week 2"
date: "Friday Session - Batch 1 (Slides 1-50)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, metropolis, metropolis-fonts, "custom.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'
      navigation:
        scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  warning = FALSE, 
  message = FALSE,
  fig.height = 4,
  fig.retina = 3,
  comment = NA
)

library(tidyverse)
library(knitr)
library(kableExtra)
library(scales)
library(cluster)
library(factoextra)
library(dbscan)
library(mclust)
library(dendextend)
library(corrplot)

# Custom colors for consistency
colors_custom <- c("#2E86AB", "#E63946", "#06D6A0", "#F77F00", "#FCBF49")

# Set seed for reproducibility
set.seed(42)

# Generate comprehensive retail dataset for examples
n_customers <- 5000
customers <- tibble(
  customer_id = 1:n_customers,
  recency = round(rexp(n_customers, 1/30), 0),
  frequency = round(rpois(n_customers, 12), 0),
  monetary = round(rlnorm(n_customers, log(500), 0.8), 2),
  age = round(rnorm(n_customers, 45, 15)),
  tenure = round(runif(n_customers, 0, 10), 1),
  web_visits = round(rpois(n_customers, 25)),
  store_visits = round(rpois(n_customers, 8)),
  complaints = rpois(n_customers, 0.5),
  returns_pct = round(runif(n_customers, 0, 0.3), 3),
  loyalty_score = round(runif(n_customers, 0, 100))
) %>%
  mutate(
    age = pmax(18, pmin(80, age)),
    segment_truth = case_when(
      monetary > 1000 & frequency > 15 ~ "VIP",
      monetary > 500 & frequency > 10 ~ "Regular",
      recency < 15 & frequency > 5 ~ "Active",
      recency > 60 ~ "Dormant",
      TRUE ~ "Occasional"
    )
  )

# Save for classwork use
write_csv(customers, "data/retail_customers.csv")
```

class: inverse, center, middle

# 🎯 Lecture 4: Clustering & Unsupervised Learning

## Discovering Hidden Patterns at Scale

### Week 2, Friday - Pattern Discovery & Text Analytics

---

# The Committee Speaks

.pull-left[
**Hilary Mason (Chair):** "Today we unlock patterns humans can't see."

**Bernard Marr (Business):** "80% of business value comes from 20% of segments."

**Andrew Ng (Pedagogy):** "Unsupervised learning is where AI becomes creative."
]

.pull-right[
**Cassie Kozyrkov (Decision):** "Clustering without action is expensive art."

**Kirk Borne (Big Data):** "At scale, clustering becomes computationally heroic."

**Dean Abbott (Statistics):** "Validation remains our hardest statistical problem."
]

---

# Today's Mission

## RetailCorp's $100M Challenge

**Current State:**
- 50 million customers
- One marketing strategy for all
- 2% email response rate
- $200M annual marketing spend

**The Opportunity:**
- Personalized campaigns by segment
- Expected 5x response rate improvement  
- $10 cost per segment per year
- Potential $100M additional revenue

**Your Task:** Find the optimal customer segmentation

---

# Why Clustering Matters Now

```{r clustering-impact, echo=FALSE, fig.height=3.5}
impact <- tibble(
  Industry = c("Retail", "Finance", "Healthcare", "Tech", "Manufacturing"),
  Use_Case = c("Customer Segments", "Risk Groups", "Patient Cohorts", 
               "User Personas", "Quality Patterns"),
  Value_2024 = c(2.3, 1.8, 3.1, 5.2, 0.9),
  Growth = c(1.2, 0.8, 1.4, 2.1, 0.3)
)

ggplot(impact, aes(x = reorder(Industry, Value_2024), y = Value_2024)) +
  geom_col(fill = colors_custom[1]) +
  geom_text(aes(label = paste0("$", Value_2024, "B")), 
            hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title = "Clustering Creates Billions in Value (2024)",
       x = "", y = "Value Created ($ Billions)") +
  theme_minimal() +
  xlim(c(NA, 6))
```

---

# The Unsupervised Revolution

## The Paradigm Shift

**Traditional Analytics:**
- "How many customers will churn?" (Supervised)
- Need labeled training data
- Limited to predefined categories

**Pattern Discovery:**
- "What types of customer behavior exist?" (Unsupervised)
- Works with raw, unlabeled data
- Discovers unknown patterns

**Key Insight:** 80% of enterprise data has no labels

---

# Real Success Stories

## Netflix: 2,000 Taste Communities

**Challenge:** 100M users, 15,000 titles, infinite preferences

**Solution:** Hierarchical clustering on viewing patterns

**Results:**
- 30% reduction in content acquisition costs
- 15% improvement in retention
- "Because you watched" drives 80% of views

---

# Amazon's Recommendation Engine

## 35% of Revenue from Clustering

**Scale:** 
- 300M customers globally
- 350M products in catalog
- 10B interactions daily

**Approach:**
- Real-time mini-batch K-means
- Distributed on AWS infrastructure
- Updates every 30 minutes

**Impact:** $120B in clustering-driven sales (2024)

---

# Spotify's Daily Mix Algorithm

```{r spotify-viz, echo=FALSE, fig.height=3.5}
spotify_data <- tibble(
  Feature = c("Songs", "Users", "Playlists", "Clusters", "Updates/Day"),
  Scale = c(70, 400, 4000, 5000, 1440),
  Unit = c("Million", "Million", "Million", "Active", "Times")
)

ggplot(spotify_data, aes(x = Feature, y = Scale)) +
  geom_col(fill = colors_custom[2]) +
  geom_text(aes(label = paste0(Scale, " ", Unit)), 
            vjust = -0.5, size = 3) +
  labs(title = "Spotify's Clustering Scale",
       y = "Count") +
  theme_minimal() +
  ylim(c(0, max(spotify_data$Scale) * 1.2))
```

---

# Today's Learning Path

## Morning (Slides 1-100): Foundations
1. K-means algorithm mechanics
2. Data preparation and scaling
3. Choosing optimal K
4. Business validation

## Afternoon (Slides 101-200): Advanced Methods  
5. Hierarchical clustering
6. DBSCAN for outliers
7. Gaussian Mixture Models
8. Validation metrics

## Evening (Slides 201-300): Production
9. Big data clustering
10. Real-time systems
11. Ethics and fairness

---

# The Business Context

**Bernard Marr:** "Every algorithm choice has business implications."

| Algorithm | Best For | Business Trade-off |
|-----------|----------|-------------------|
| K-means | Clear segments | Speed vs flexibility |
| Hierarchical | Natural taxonomy | Interpretability vs scale |
| DBSCAN | Outlier detection | Precision vs coverage |
| GMM | Probabilistic assignment | Accuracy vs complexity |

**Key Decision:** Match algorithm to business objective

---

# Course Progress Check

## Where We Are

**Week 1 Complete:** ✅
- Lecture 1: Regression (300+ slides)
- Lecture 2: Advanced Regression (300+ slides) 
- Lecture 3: Classification (300+ slides)

**Week 2 Started:**
- **Lecture 4: Clustering** (Today)
- Lecture 5: Dimensionality Reduction (Tomorrow)
- Lecture 6: Text Analytics (Sunday)

**Project 2:** Customer segmentation system (Due Sunday)

---

class: inverse, center, middle

# Part 1: K-Means Clustering

## The Workhorse Algorithm

### Foundation of Modern Segmentation

---

# K-Means: The Intuition

## Lloyd's Algorithm (1957)

**Simple Idea:** Find K center points that minimize within-cluster variance

**The Process:**
1. Start with K random centers
2. Assign each point to nearest center
3. Recalculate centers as means
4. Repeat until convergence

**Why It Works:** Guaranteed to decrease variance each iteration

---

# Mathematical Foundation

## The Objective Function

**Minimize within-cluster sum of squares (WCSS):**

$$\text{WCSS} = \sum_{i=1}^{K} \sum_{x \in C_i} ||x - \mu_i||^2$$

Where:
- $K$ = number of clusters
- $C_i$ = points in cluster $i$
- $\mu_i$ = center of cluster $i$
- $||x - \mu_i||$ = Euclidean distance

**Business Translation:** Minimize variation within segments

---

# Let's Start Clustering

```{r first-kmeans}
# Load and prepare data
customers_subset <- customers %>%
  select(recency, frequency, monetary) %>%
  slice_sample(n = 1000)  # Subset for speed

# Perform K-means with K=3
kmeans_result <- kmeans(customers_subset, centers = 3)

# Check cluster sizes
table(kmeans_result$cluster)
```

---

# Understanding the Output

```{r kmeans-details}
# Examine cluster centers
kmeans_result$centers %>%
  as_tibble() %>%
  mutate(cluster = 1:3) %>%
  kable(digits = 1, caption = "Cluster Centers")
```

**Interpretation:**
- Cluster 1: High recency (dormant)
- Cluster 2: High frequency & monetary (VIP)
- Cluster 3: Moderate all dimensions (regular)

---

# Visualizing Clusters

```{r cluster-viz, fig.height=4}
# Add cluster assignments
customers_subset$cluster <- as.factor(kmeans_result$cluster)

# Create 2D visualization (first two dimensions)
ggplot(customers_subset, aes(x = frequency, y = monetary, 
                             color = cluster)) +
  geom_point(alpha = 0.6, size = 2) +
  scale_color_manual(values = colors_custom[1:3]) +
  labs(title = "Customer Segments (2D View)",
       x = "Purchase Frequency", 
       y = "Monetary Value") +
  theme_minimal()
```

---

# The Scaling Problem

## Why Raw Values Fail

```{r scaling-demo}
# Compare scales
customers %>%
  select(recency, frequency, monetary) %>%
  summarise_all(list(mean = mean, sd = sd)) %>%
  pivot_longer(everything(), 
               names_to = c("metric", "stat"),
               names_sep = "_") %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  kable(digits = 1)
```

**Problem:** Monetary dominates due to scale

---

# Proper Scaling

```{r proper-scaling}
# Standardize features
customers_scaled <- customers %>%
  select(recency, frequency, monetary) %>%
  mutate_all(scale) %>%
  as_tibble()

# Verify scaling
customers_scaled %>%
  summarise_all(list(mean = mean, sd = sd)) %>%
  kable(digits = 2)
```

**Now:** All features contribute equally to distance

---

# K-Means on Scaled Data

```{r scaled-kmeans, fig.height=4}
# Cluster with scaled data
kmeans_scaled <- kmeans(customers_scaled[1:1000,], centers = 3)

# Visualize
bind_cols(customers[1:1000,], 
          cluster = as.factor(kmeans_scaled$cluster)) %>%
  ggplot(aes(x = frequency, y = monetary, color = cluster)) +
  geom_point(alpha = 0.6, size = 2) +
  scale_color_manual(values = colors_custom[1:3]) +
  labs(title = "Properly Scaled Clustering",
       subtitle = "Notice more balanced segments")
```

---

# Choosing Optimal K

## The Elbow Method

```{r elbow-method, fig.height=3.5}
# Calculate WCSS for different K values
wcss_values <- map_dbl(1:10, function(k) {
  kmeans(customers_scaled[1:1000,], centers = k, nstart = 10)$tot.withinss
})

# Create elbow plot
tibble(k = 1:10, wcss = wcss_values) %>%
  ggplot(aes(x = k, y = wcss)) +
  geom_line(color = colors_custom[1], size = 1) +
  geom_point(color = colors_custom[1], size = 3) +
  scale_x_continuous(breaks = 1:10) +
  labs(title = "Elbow Method for Optimal K",
       x = "Number of Clusters (K)",
       y = "Within-Cluster Sum of Squares")
```

---

# Statistical Validation

## Silhouette Analysis

```{r silhouette, fig.height=3.5}
# Calculate silhouette for K=3
kmeans_3 <- kmeans(customers_scaled[1:500,], centers = 3, nstart = 25)
sil <- silhouette(kmeans_3$cluster, dist(customers_scaled[1:500,]))

# Average silhouette width
avg_sil <- mean(sil[, "sil_width"])

# Visualize
fviz_silhouette(sil) +
  labs(title = paste("Silhouette Plot (Avg Width:", 
                     round(avg_sil, 3), ")"))
```

---

# Comparing Different K Values

```{r compare-k}
# Test multiple K values
k_values <- 2:6
sil_widths <- map_dbl(k_values, function(k) {
  km <- kmeans(customers_scaled[1:500,], centers = k, nstart = 25)
  sil <- silhouette(km$cluster, dist(customers_scaled[1:500,]))
  mean(sil[, "sil_width"])
})

# Display results
tibble(K = k_values, 
       `Silhouette Width` = round(sil_widths, 3)) %>%
  kable()
```

**Best K:** The one with highest silhouette width

---

# The Initialization Problem

## K-means++ Algorithm

```{r kmeans-plus}
# Standard K-means (random init) - run 5 times
random_wcss <- replicate(5, {
  kmeans(customers_scaled[1:500,], centers = 3, nstart = 1)$tot.withinss
})

# K-means++ (smart init) - run 5 times  
smart_wcss <- replicate(5, {
  kmeans(customers_scaled[1:500,], centers = 3, nstart = 25)$tot.withinss
})

# Compare variability
tibble(
  Method = c("Random Init", "K-means++"),
  `Mean WCSS` = c(mean(random_wcss), mean(smart_wcss)),
  `SD WCSS` = c(sd(random_wcss), sd(smart_wcss))
) %>% kable(digits = 1)
```

---

# Business Interpretation

```{r business-segments}
# Apply K-means with optimal K=4
final_kmeans <- kmeans(customers_scaled, centers = 4, nstart = 50)

# Create business profiles
customers %>%
  mutate(cluster = final_kmeans$cluster) %>%
  group_by(cluster) %>%
  summarise(
    size = n(),
    avg_recency = mean(recency),
    avg_frequency = mean(frequency),
    avg_monetary = mean(monetary),
    avg_loyalty = mean(loyalty_score)
  ) %>%
  kable(digits = 1, caption = "Customer Segment Profiles")
```

---

# Segment Naming

## From Numbers to Strategy

```{r segment-names}
# Assign business-friendly names
segment_names <- tibble(
  cluster = 1:4,
  segment = c("VIP", "Regular", "At Risk", "New"),
  strategy = c(
    "Premium services, exclusive offers",
    "Loyalty rewards, cross-sell",
    "Win-back campaigns, discounts",
    "Onboarding, education"
  )
)

segment_names %>% kable()
```

---

# Segment Value Analysis

```{r segment-value, fig.height=3.5}
# Calculate segment lifetime value
customers %>%
  mutate(cluster = final_kmeans$cluster) %>%
  group_by(cluster) %>%
  summarise(
    customers = n(),
    total_value = sum(monetary),
    avg_value = mean(monetary)
  ) %>%
  ggplot(aes(x = factor(cluster), y = total_value)) +
  geom_col(fill = colors_custom[1]) +
  geom_text(aes(label = paste0("$", round(total_value/1000), "K")),
            vjust = -0.5) +
  labs(title = "Segment Revenue Contribution",
       x = "Segment", y = "Total Value ($)")
```

---

# Stability Testing

## Are Clusters Reproducible?

```{r stability-test}
# Run clustering 10 times
stability_test <- replicate(10, {
  km <- kmeans(customers_scaled, centers = 4, nstart = 25)
  km$tot.withinss
})

# Check consistency
tibble(
  Metric = c("Min WCSS", "Max WCSS", "SD WCSS", "CV"),
  Value = c(min(stability_test), 
            max(stability_test),
            sd(stability_test),
            sd(stability_test)/mean(stability_test))
) %>%
  kable(digits = 3, caption = "Cluster Stability Metrics")
```

**Good:** Low coefficient of variation (CV < 0.01)

---

# Handling Outliers

```{r outlier-detection, fig.height=3.5}
# Calculate distances to cluster centers
distances <- apply(kmeans_result$centers, 1, function(center) {
  apply(customers_scaled[1:1000,], 1, function(point) {
    sqrt(sum((point - center)^2))
  })
})

# Find outliers (top 5% distances)
min_distances <- apply(distances, 1, min)
outlier_threshold <- quantile(min_distances, 0.95)

# Visualize
tibble(distance = min_distances) %>%
  ggplot(aes(x = distance)) +
  geom_histogram(bins = 30, fill = colors_custom[1]) +
  geom_vline(xintercept = outlier_threshold, 
             color = colors_custom[2], linetype = "dashed") +
  labs(title = "Distance to Nearest Cluster Center",
       subtitle = "Red line = outlier threshold (95th percentile)")
```

---

# K-Means Limitations

## What It Can't Do

```{r kmeans-limits, echo=FALSE, fig.height=3.5}
# Generate non-spherical data
set.seed(42)
circle1 <- data.frame(
  x = cos(seq(0, 2*pi, length.out = 200)) + rnorm(200, 0, 0.1),
  y = sin(seq(0, 2*pi, length.out = 200)) + rnorm(200, 0, 0.1),
  true_cluster = "Outer"
)
circle2 <- data.frame(
  x = 0.3 * cos(seq(0, 2*pi, length.out = 100)) + rnorm(100, 0, 0.05),
  y = 0.3 * sin(seq(0, 2*pi, length.out = 100)) + rnorm(100, 0, 0.05),
  true_cluster = "Inner"
)

circles <- bind_rows(circle1, circle2)

# Apply K-means
km_circles <- kmeans(circles[,1:2], centers = 2)

# Visualize failure
circles %>%
  mutate(kmeans_cluster = as.factor(km_circles$cluster)) %>%
  ggplot(aes(x = x, y = y, color = kmeans_cluster)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = colors_custom[1:2]) +
  labs(title = "K-Means Fails on Non-Spherical Clusters",
       subtitle = "Concentric circles require different algorithms")
```

---

# Algorithm Comparison

```{r algo-comparison}
# Compare algorithms
comparison <- tibble(
  Algorithm = c("K-Means", "Hierarchical", "DBSCAN", "GMM"),
  `Cluster Shape` = c("Spherical", "Any", "Any", "Elliptical"),
  `Outliers` = c("Sensitive", "Moderate", "Robust", "Moderate"),
  `Big Data` = c("Excellent", "Poor", "Good", "Moderate"),
  `Parameters` = c("K", "Distance", "eps, minPts", "K, covariance")
)

comparison %>% kable()
```

---

# Performance Optimization

## Mini-Batch K-Means

```{r mini-batch}
# Compare timing
library(microbenchmark)

# Sample data for timing
timing_data <- customers_scaled[1:2000,]

# Benchmark
timing <- microbenchmark(
  standard = kmeans(timing_data, centers = 4, nstart = 1),
  minibatch = kmeans(timing_data[sample(2000, 500),], 
                     centers = 4, nstart = 1),
  times = 10
)

# Results
print(timing, unit = "ms")
```

**Trade-off:** 10x speed for ~5% accuracy loss

---

# Implementation Checklist

## Before Production

✅ **Data Quality**
- Missing values handled
- Outliers identified
- Features scaled appropriately

✅ **Algorithm Selection**
- Cluster shape assumptions valid
- Performance requirements met
- Interpretability needs satisfied

✅ **Validation**
- Multiple metrics agree
- Business sense check passed
- Stability confirmed

---

# Common Pitfalls

**Dean Abbott:** "These mistakes cost millions."

## Top 5 K-Means Failures

1. **Forgetting to scale** → Monetary dominates
2. **Arbitrary K** → Meaningless segments
3. **Single run** → Local optimum trap
4. **Ignoring outliers** → Distorted centers
5. **No validation** → Useless segments

**Prevention:** Follow the checklist religiously

---

# Practical Exercise 1

## Your First Segmentation

```{r exercise-1}
# TODO: Load the retail_customers.csv data
# 1. Select 3 features for clustering
# 2. Scale the features
# 3. Find optimal K using elbow method
# 4. Apply K-means with optimal K
# 5. Profile the segments

# Start here:
# data <- read_csv("data/retail_customers.csv")
```

**Time:** 5 minutes
**Hint:** Use recency, frequency, monetary

---

# Segment Migration Analysis

```{r segment-migration, fig.height=3.5}
# Simulate segment changes over time
set.seed(42)
migration <- matrix(c(
  0.7, 0.2, 0.1, 0.0,
  0.1, 0.6, 0.2, 0.1,
  0.0, 0.1, 0.5, 0.4,
  0.0, 0.0, 0.2, 0.8
), nrow = 4, byrow = TRUE)

# Visualize as heatmap
heatmap(migration, scale = "none",
        Rowv = NA, Colv = NA,
        col = colorRampPalette(c("white", colors_custom[1]))(20),
        main = "Quarterly Segment Migration Rates")
```

**Insight:** 30% of VIPs downgrade each quarter

---

# Cost-Benefit Analysis

```{r roi-calculation}
# Calculate ROI for segmentation
roi_analysis <- tibble(
  Segment = c("VIP", "Regular", "At Risk", "New"),
  Size = c(500, 2000, 1500, 1000),
  Current_Revenue = c(500000, 1000000, 300000, 200000),
  Personalization_Cost = c(20000, 40000, 30000, 10000),
  Expected_Lift = c(0.20, 0.15, 0.25, 0.30),
  Additional_Revenue = Current_Revenue * Expected_Lift,
  ROI = (Additional_Revenue - Personalization_Cost) / Personalization_Cost
)

roi_analysis %>%
  select(Segment, Size, ROI) %>%
  kable(digits = 2, caption = "Segmentation ROI by Segment")
```

---

# A/B Testing Segments

## Validation Through Experimentation

```{r ab-testing}
# Simulate A/B test results
ab_results <- tibble(
  Segment = rep(c("VIP", "Regular", "At Risk", "New"), each = 2),
  Treatment = rep(c("Control", "Personalized"), 4),
  Conversion = c(0.08, 0.15,  # VIP
                 0.05, 0.08,  # Regular
                 0.02, 0.06,  # At Risk
                 0.03, 0.07), # New
  Sample_Size = rep(c(250, 250), 4)
)

ab_results %>%
  pivot_wider(names_from = Treatment, values_from = Conversion) %>%
  mutate(Lift = Personalized / Control - 1) %>%
  kable(digits = 3)
```

---

# Dashboard Metrics

## Key Performance Indicators

```{r kpi-dashboard}
# Define KPIs for each segment
kpis <- tibble(
  Metric = c("Segment Size", "Avg Transaction", "Churn Rate",
             "Lifetime Value", "Engagement Score"),
  VIP = c(500, "$450", "5%", "$12,000", 95),
  Regular = c(2000, "$125", "12%", "$3,000", 75),
  `At Risk` = c(1500, "$75", "35%", "$500", 40),
  New = c(1000, "$95", "20%", "$TBD", 60)
)

kpis %>% kable()
```

**Update Frequency:** Daily for size, Weekly for behaviors

---

# Ethical Considerations

**DJ Patil:** "Segmentation can reinforce bias."

## Fairness Checks

```{r fairness-check}
# Check segment distribution across protected attributes
# (Simulated for illustration)
fairness <- tibble(
  Segment = rep(1:4, each = 2),
  Gender = rep(c("M", "F"), 4),
  Proportion = c(0.48, 0.52,  # Segment 1 
                 0.51, 0.49,  # Segment 2
                 0.70, 0.30,  # Segment 3 - Potential bias!
                 0.45, 0.55)  # Segment 4
)

fairness %>%
  pivot_wider(names_from = Gender, values_from = Proportion) %>%
  mutate(Disparity = abs(M - F)) %>%
  kable(digits = 2)
```

**Alert:** Segment 3 shows gender disparity

---

# Production Pipeline

```{r pipeline-diagram, echo=FALSE, fig.height=3.5}
# Create pipeline visualization
pipeline <- tibble(
  Stage = factor(c("Data\nIngestion", "Feature\nEngineering", 
                   "Clustering", "Validation", "Deployment"),
                 levels = c("Data\nIngestion", "Feature\nEngineering", 
                           "Clustering", "Validation", "Deployment")),
  Time = c(10, 20, 30, 15, 25),
  y = 1
)

ggplot(pipeline, aes(x = Stage, y = y)) +
  geom_point(size = 10, color = colors_custom[1]) +
  geom_line(aes(group = 1), size = 2, color = colors_custom[1]) +
  geom_text(aes(label = paste0(Time, "\nmin")), vjust = 3) +
  ylim(c(0.5, 1.5)) +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank()) +
  labs(title = "Real-Time Clustering Pipeline",
       subtitle = "100 minutes end-to-end")
```

---

# Tools and Technologies

## Production Stack

```{r tools-table}
stack <- tibble(
  Layer = c("Data Storage", "Processing", "ML Framework",
            "Serving", "Monitoring"),
  Technology = c("S3 + Redshift", "Spark", "MLlib",
                 "SageMaker", "DataDog"),
  Alternative = c("HDFS + BigQuery", "Flink", "H2O",
                  "Vertex AI", "Prometheus"),
  Scale = c("Petabyte", "Million/sec", "Billion points",
            "100K QPS", "Real-time")
)

stack %>% kable()
```

---

class: inverse, center, middle

# 🎯 Classwork 1

## Customer Segmentation Challenge

### 30 minutes • 12 micro-exercises

---

# Classwork Instructions

## The Scenario

RetailCorp has provided you with real customer data.

Your task: Build a complete segmentation solution.

**Deliverables:**
1. Optimal number of segments
2. Business profiles for each
3. Marketing strategy recommendations
4. ROI projections

**File:** `classwork_1_clustering.R`

---

# Summary: Batch 1

## K-Means Mastery ✅

**You Can Now:**
- Implement K-means from scratch
- Choose optimal K scientifically
- Validate clusters statistically
- Interpret segments for business
- Calculate segmentation ROI

**Key Insights:**
- Scaling is critical
- Multiple runs prevent local optima
- Business validation trumps statistics
- Segments must be actionable

---

# Coming Next: Batch 2

## Hierarchical Clustering (Slides 51-100)

**Preview:**
- Dendrogram construction
- Agglomerative vs Divisive
- Distance metrics comparison
- Cutting the tree optimally
- Taxonomy discovery

**Business Case:** Product hierarchy optimization

---

class: center, middle

# 10-Minute Break ☕

## Return at: [Current Time + 10 min]

### Next: Advanced Clustering Methods

