---
title: "Classwork 7: Production Deployment Planning"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

# Setup

```{r libraries}
library(tidyverse)
library(glmnet)
library(kableExtra)

# Load production data
production_data <- read_csv("production_simulation.csv")

# Load pre-trained model
ridge_model <- readRDS("production_model.rds")

# Load training statistics for comparison
training_stats <- readRDS("training_stats.rds")

# Set seed for reproducibility
set.seed(123)
```

---

# Part 1: Production Performance Evaluation

## Load and Score Production Data

```{r score-production}
# Prepare feature matrix
# TODO: Extract features and prepare matrix format

# Generate predictions
# TODO: Use ridge_model to score production data

# Calculate performance metrics
# TODO: Compute RMSE, MAE, R²
```

## Overall Performance Comparison

```{r performance-table}
# TODO: Create comparison table
# Columns: Metric | Validation | Production | Change (%)
# Metrics: RMSE, MAE, R²
```

**Performance Summary:**

[Write 2-3 sentences describing overall performance degradation. Is the model still acceptable? Which metrics degraded most?]

## Segment Analysis

```{r segment-performance}
# TODO: Calculate RMSE by product category

# TODO: Calculate RMSE by price tier (quantile-based groups)

# TODO: Calculate RMSE by brand (if available)

# TODO: Identify worst-performing segments
```

```{r segment-visualization}
# TODO: Create bar plot showing RMSE by segment
# Compare to overall production RMSE
```

**Segment Findings:**

[Write 3-4 sentences identifying which segments show severe degradation and potential explanations.]

## Error Distribution Analysis

```{r error-analysis}
# TODO: Calculate prediction errors

# TODO: Compute error percentiles (50th, 75th, 90th, 95th, 99th)

# TODO: Identify products with extreme errors (>$50)
```

**Degradation Assessment:**

[Write 2-3 sentences on whether degradation is uniform or concentrated. What does this tell you about deployment risk?]

---

# Part 2: Feature Drift Analysis

## Continuous Feature Drift

```{r feature-drift}
# TODO: For each continuous feature, calculate:
# - Mean in training vs production
# - SD in training vs production  
# - Percentage change in mean
# - Percentage change in SD

# TODO: Flag features with >20% mean shift or >30% SD change
```

```{r drift-visualization}
# TODO: Create scatter plot showing mean shift (x) vs SD change (y)
# Color points by severity: green (acceptable), yellow (moderate), red (severe)
# Add reference lines at 20% and 30% thresholds
```

**Drift Summary:**

[Write 2-3 sentences listing the most drifted features and their shift magnitudes.]

## Missing Value Analysis

```{r missing-analysis}
# TODO: Calculate missing rate by feature in training data

# TODO: Calculate missing rate by feature in production data

# TODO: Compute increase in missing rates

# TODO: Flag features with >10% missing rate increase
```

```{r missing-visualization}
# TODO: Create comparison plot showing training vs production missing rates
# Highlight features exceeding 15% total missing rate
```

**Missing Value Findings:**

[Write 2-3 sentences on which features have elevated missing rates and whether this correlates with performance degradation.]

## High-Risk Features

```{r risk-features}
# TODO: Create summary table of high-risk features
# Columns: Feature | Mean Shift (%) | SD Change (%) | Missing Increase (%) | Risk Level
# Include only features exceeding at least one threshold
```

**Risk Assessment:**

[Write 3-4 sentences on how many features show problematic drift and whether this explains performance degradation. Should retraining be triggered?]

---

# Part 3: Monitoring Dashboard Design

## Dashboard Metrics Specification

```{r dashboard-metrics}
# TODO: Create metrics tracking table
# Columns: Metric | Current Value | Threshold (Yellow) | Threshold (Red) | Alert Status
# Include: RMSE, MAE, Mean Feature Drift, Max Feature Drift, Missing Rate, Prediction Volume

# TODO: Calculate current alert status for each metric
```

## Time-Series Monitoring (Simulated)

```{r timeseries-simulation}
# TODO: Simulate 30 days of RMSE values showing gradual drift
# Days 1-10: RMSE around $8.5
# Days 11-20: RMSE around $9.2  
# Days 21-30: RMSE around $10.5

# TODO: Plot time series with threshold lines for yellow ($9.50) and red ($10.00) alerts
```

## Feature Drift Dashboard

```{r drift-dashboard}
# TODO: Create heatmap showing feature drift over time (simulated)
# Rows: Top 20 features by importance
# Columns: Days (1-30)
# Color: Drift magnitude
```

**Dashboard Design Summary:**

[Write 2-3 sentences describing how this dashboard enables proactive monitoring and early detection of issues.]

---

# Part 4: Retraining Strategy

## Retraining Schedule

```{r retraining-schedule}
# TODO: Create retraining options comparison table
# Columns: Schedule | Frequency | Annual Cost | Pros | Cons
# Rows: Weekly, Bi-weekly, Monthly, Triggered-only
```

**Recommended Schedule:**

[Write 2-3 sentences justifying your recommended retraining frequency based on drift severity and costs.]

## Retraining Triggers

```{r trigger-specification}
# TODO: Create trigger specification table
# Columns: Trigger Condition | Threshold | Action | Priority
# Examples: RMSE degradation >15%, Feature drift >30%, Missing rate >20%
```

## Retraining Cost Analysis

```{r cost-analysis}
# TODO: Calculate annual costs for different strategies
# Include: Data scientist time, computational resources, testing/validation
# Consider: Scheduled retraining + estimated triggered retraining

# Weekly: 52 retrains/year
# Monthly: 12 retrains/year  
# Triggered: Estimated based on current drift rate
```

**Cost-Benefit Assessment:**

[Write 2-3 sentences weighing retraining costs against risk of degraded model performance. What's optimal?]

---

# Part 5: Risk Mitigation Plan

## Risk Identification Matrix

```{r risk-matrix}
# TODO: Create risk matrix table
# Columns: Risk | Likelihood | Impact | Severity Score | Mitigation Strategy
# Risks: Data pipeline failure, Feature drift, Model bugs, Infrastructure outage, 
#        Data quality issues, Correlation structure changes
```

## Rollback Procedures

**Rollback Decision Criteria:**

[Write a bulleted list of specific conditions that would trigger model rollback, with quantitative thresholds.]

**Rollback Process:**

[Write 4-5 sentences describing step-by-step rollback procedures including who approves, how quickly it occurs, and how to communicate to stakeholders.]

## Incident Response Plan

```{r incident-response}
# TODO: Create incident response table
# Columns: Alert Level | Response Time | Responsible Team | Actions | Escalation Path
# Levels: Green (normal), Yellow (investigate), Red (immediate action)
```

**Communication Protocols:**

[Write 2-3 sentences on how alerts are communicated and who is notified at each severity level.]

---

# Part 6: Deployment Documentation

## Deployment Summary Memo

**TO:** VP of Engineering  
**FROM:** [Your Name], Senior Data Scientist  
**DATE:** `r Sys.Date()`  
**RE:** Ridge Regression Pricing Model - Production Deployment Plan

### Model Specifications

[Write 1 paragraph describing model type, features, training approach, and performance benchmarks.]

### Production Performance

[Write 1 paragraph summarizing production evaluation findings, including current RMSE, degradation level, and segment-specific issues.]

### Monitoring Plan

[Write 1 paragraph describing monitoring metrics, alert thresholds, dashboard design, and review frequency.]

### Retraining Strategy

[Write 1 paragraph specifying retraining schedule, triggers, costs, and expected maintenance requirements.]

### Risk Mitigation

[Write 1 paragraph outlining key risks, mitigation strategies, rollback procedures, and incident response protocols.]

### Deployment Recommendation

[Write 2-3 sentences with your deployment recommendation: proceed, proceed with conditions, or delay. Justify based on evidence from your analysis.]

---

## Critical Feature Requirements

```{r feature-requirements}
# TODO: Create feature requirements table
# Top 10 features by coefficient magnitude
# Columns: Feature | Coefficient | Expected Range | Acceptable Missing (%) | Latency Budget (ms)
```

## Maintenance Calendar

```{r maintenance-calendar}
# TODO: Create maintenance schedule table
# Columns: Activity | Frequency | Responsible Party | Estimated Duration
# Include: Performance reviews, retraining, audits, architecture reviews
```

---

# Summary and Key Learnings

## What I Learned

1. Production deployment requires...

2. Feature drift can...

3. Monitoring infrastructure must...

## Critical Success Factors

[Write 3-4 sentences on what determines successful production deployment of regularized models.]

## Recommendations for Future Deployments

[Write 2-3 sentences on lessons learned that would improve future model deployments.]

---

# Session Info

```{r session-info}
sessionInfo()
```
