---
title: "Classwork 4: Elastic Net Practice"
subtitle: "Combining Ridge and LASSO"
author: "Your Name Here"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

library(tidyverse)
library(glmnet)
library(caret)
library(knitr)

theme_set(theme_minimal(base_size = 14))
```

---

# Introduction

**Scenario:** You're continuing your work at **HomeSmart Realty**. After implementing Ridge and LASSO, your manager asks:

> "Can we get the best of both worlds? I want feature selection like LASSO, but I'm worried about the correlation between square footage and lot size making LASSO unstable."

**Your task:** Implement Elastic Net to balance Ridge and LASSO benefits.

---

# Load the Data

```{r load_data}
# Use the same housing data from Classwork 3
set.seed(42)

# Generate synthetic housing data
n <- 500

housing_data <- tibble(
  house_id = 1:n,
  square_feet = rnorm(n, 2000, 500),
  bedrooms = sample(2:6, n, replace = TRUE),
  bathrooms = sample(1:4, n, replace = TRUE),
  lot_size = rnorm(n, 8000, 2000),
  age_years = sample(1:50, n, replace = TRUE),
  garage_spaces = sample(0:3, n, replace = TRUE),
  school_rating = rnorm(n, 7, 1.5),
  crime_index = rnorm(n, 50, 15),
  distance_downtown = rnorm(n, 5, 2),
  nearby_parks = sample(0:5, n, replace = TRUE),
  hoa_fees = rnorm(n, 200, 100),
  property_tax = rnorm(n, 3000, 800)
) %>%
  mutate(
    price = 100000 + 
      150 * square_feet + 
      20000 * bedrooms +
      15000 * bathrooms +
      10 * lot_size +
      -2000 * age_years +
      10000 * garage_spaces +
      8000 * school_rating +
      -500 * crime_index +
      -3000 * distance_downtown +
      5000 * nearby_parks +
      rnorm(n, 0, 30000)
  ) %>%
  mutate(price = pmax(price, 50000))

# Prepare train/test split and matrices
set.seed(123)
train_idx <- createDataPartition(housing_data$price, p = 0.8, list = FALSE)

X_train <- model.matrix(price ~ . - house_id, data = housing_data[train_idx, ])[, -1]
y_train <- housing_data$price[train_idx]

X_test <- model.matrix(price ~ . - house_id, data = housing_data[-train_idx, ])[, -1]
y_test <- housing_data$price[-train_idx]

cat("Data ready! X_train dimensions:", dim(X_train), "\n")
```

---

# Exercise 1: Fit Basic Elastic Net (α = 0.5)

**Task:** Fit an Elastic Net model with alpha = 0.5 and lambda = 1000.

```{r ex1}
# YOUR CODE HERE
# Hint: Use glmnet() with alpha = 0.5




# Print coefficients


# How many features were selected (non-zero)?
```

**Question:** How many features did Elastic Net select with this lambda?

---

# Exercise 2: Cross-Validation for Elastic Net

**Task:** Use cv.glmnet() to find the optimal lambda for Elastic Net (α = 0.5).

```{r ex2}
# YOUR CODE HERE
set.seed(123)
# Hint: Use cv.glmnet() with alpha = 0.5




# Print optimal lambda values


# Plot cross-validation results
```

**Question:** What are lambda.min and lambda.1se? Which should you use?

---

# Exercise 3: Fit Final Elastic Net Model

**Task:** Fit the final Elastic Net model using lambda.1se.

```{r ex3}
# YOUR CODE HERE
# Use the optimal lambda from Exercise 2




# Extract and print coefficients


# Count selected features
```

**Question:** How many features did Elastic Net select? List them.

---

# Exercise 4: Compare to Ridge and LASSO

**Task:** Fit Ridge and LASSO models (from Classwork 3) and compare coefficients.

```{r ex4}
# YOUR CODE HERE
# Fit Ridge (alpha = 0)
set.seed(123)



# Fit LASSO (alpha = 1)
set.seed(123)



# Compare coefficients side-by-side


# Create comparison visualization
```

**Question:** Which features are selected by all three methods? Which only by Elastic Net?

---

# Exercise 5: Performance Comparison

**Task:** Calculate test RMSE for OLS, Ridge, LASSO, and Elastic Net.

```{r ex5}
# YOUR CODE HERE
# Calculate test RMSE for all four methods




# Create comparison table


# Visualize performance
```

**Question:** Which method performs best on test data? By how much?

---

# Exercise 6: Tune Alpha Parameter

**Task:** Test different alpha values (0, 0.2, 0.4, 0.6, 0.8, 1.0) to find the optimal mix.

```{r ex6}
# YOUR CODE HERE
# Test multiple alpha values
alpha_values <- c(0, 0.2, 0.4, 0.6, 0.8, 1.0)




# For each alpha:
# - Fit with cross-validation
# - Get test RMSE
# - Count features selected


# Create results table


# Plot test RMSE vs alpha
```

**Question:** Which alpha gives the best test performance? How many features does it select?

---

# Exercise 7: Stability Analysis

**Task:** Test the stability of feature selection across different data samples.

```{r ex7}
# YOUR CODE HERE
# Fit Elastic Net on 5 different bootstrap samples




# For each sample:
# - Resample training data (90%)
# - Fit Elastic Net with optimal alpha
# - Record which features are selected


# Summarize: How many times was each feature selected?


# Visualize stability
```

**Question:** Which features are consistently selected? Which are unstable?

---

# Exercise 8: Business Recommendation

**Task:** Create a final recommendation for HomeSmart Realty comparing all methods.

```{r ex8}
# YOUR CODE HERE
# Create comprehensive comparison table




# Include:
# - Method name
# - Alpha value
# - Number of features
# - Test RMSE
# - Interpretability rating
# - Recommendation (Yes/No)


# Write your recommendation as a comment
```

**Question:** Which method would you recommend to HomeSmart Realty? Why?

**Write your answer here:**
```
Based on my analysis, I recommend [METHOD] because...

Reasons:
1. 
2. 
3. 

Key features identified:
- 
- 
- 
```

---

# Bonus Challenge 1: Coefficient Paths

**Task:** Plot coefficient paths for Elastic Net to see how coefficients change with lambda.

```{r bonus1, eval=FALSE}
# YOUR CODE HERE
# Fit Elastic Net across many lambda values




# Plot coefficient paths


# Add vertical line at optimal lambda
```

**Question:** Which features enter the model first (at high lambda)?

---

# Bonus Challenge 2: 2D Grid Search

**Task:** Perform a 2D grid search over both alpha and lambda simultaneously.

```{r bonus2, eval=FALSE}
# YOUR CODE HERE
# Create grid of alpha and lambda values




# For each combination:
# - Fit model
# - Calculate test RMSE


# Find best combination


# Create heatmap of performance
```

**Question:** What's the optimal alpha-lambda combination?

---

# Bonus Challenge 3: Feature Correlation Analysis

**Task:** Analyze which correlated features are kept together by Elastic Net vs dropped by LASSO.

```{r bonus3, eval=FALSE}
# YOUR CODE HERE
# Calculate correlation matrix




# Identify highly correlated pairs (r > 0.7)


# Check which method keeps both vs picks one


# Visualize the grouping effect
```

**Question:** Does Elastic Net keep correlated features together better than LASSO?

---

# Summary Questions

**Answer these conceptual questions:**

1. **What does alpha control in Elastic Net?**
   - Your answer:

2. **When would you prefer Elastic Net over Ridge?**
   - Your answer:

3. **When would you prefer Elastic Net over LASSO?**
   - Your answer:

4. **What is the "grouping effect" in Elastic Net?**
   - Your answer:

5. **How do you choose optimal alpha?**
   - Your answer:

---

# Self-Check: Did You...

- [ ] Fit Elastic Net with alpha = 0.5
- [ ] Use cross-validation to find optimal lambda
- [ ] Compare Elastic Net to Ridge and LASSO
- [ ] Tune the alpha parameter
- [ ] Test feature selection stability
- [ ] Calculate test RMSE for all methods
- [ ] Make a business recommendation
- [ ] Answer all questions

---

# Submission Instructions

**To submit your work:**

1. Complete all 8 exercises
2. Answer all questions in comments or text
3. Run all code chunks to verify they work
4. Knit this document to HTML
5. Submit the HTML file to Canvas

**Grading Criteria:**
- Code runs without errors (40%)
- Correct implementation (40%)
- Quality of answers and insights (20%)

**Total Points:** 100

---

# Expected Results

**Your Elastic Net model should:**
- Select approximately 8-10 features
- Achieve test RMSE around $30,000-35,000
- Perform similarly to Ridge and LASSO
- Show more stability than LASSO alone
- Provide clear business insights

**If your results differ significantly, check:**
- Random seed is set correctly
- Data preparation matches instructions
- Alpha and lambda values are appropriate

---

**Good luck! 🚀**
