---
title: "Classwork 3: Advanced Interpretation"
author: "Your Name"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(car)
library(lm.beta)
library(gridExtra)

theme_set(theme_minimal(base_size = 12))
course_colors <- c('#2E86AB', '#A23B72', '#F18F01', '#C73E1D', '#6A994E', '#BC4B51')
```

# Load Data and Fit Model
```{r}
# Load store performance data
store_data <- read_csv('data/store_performance.csv')

# Fit full model
model_full <- lm(profit ~ foot_traffic + employee_training_hours + 
                   local_income + store_age + parking_spaces, 
                 data = store_data)

summary(model_full)
```

---

# Task 1: Partial Regression Plots

## Method 1: Manual Calculation (Foot Traffic)
```{r}
# TODO: Create partial regression plot for foot_traffic

# Step 1: Regress profit on all predictors EXCEPT foot_traffic
model_no_traffic <- lm(___ ~ ___, data = store_data)
resid_profit <- residuals(___)

# Step 2: Regress foot_traffic on all other predictors
model_traffic_on_others <- lm(___ ~ ___, data = store_data)
resid_traffic <- residuals(___)

# Step 3: Plot
tibble(resid_traffic = resid_traffic,
       resid_profit = resid_profit) %>%
  ggplot(aes(x = resid_traffic, y = resid_profit)) +
  geom_point(alpha = 0.6, color = course_colors[1]) +
  geom_smooth(method = "lm", se = TRUE, color = course_colors[3]) +
  labs(title = "Partial Regression: Foot Traffic",
       x = "Foot Traffic (residualized)",
       y = "Profit (residualized)")

# Verify slope matches coefficient
cat("Partial regression slope:", coef(lm(resid_profit ~ resid_traffic))[2], "\n")
cat("Full model coefficient:", coef(model_full)["foot_traffic"], "\n")
```

**Interpretation:**
```
TODO: One sentence explaining what this plot shows
```

## Method 2: Using avPlots() for All Predictors
```{r fig.height=6}
# TODO: Use car::avPlots() to generate all partial regression plots at once
avPlots(___)
```

**Strongest relationship:**
```
TODO: Which predictor shows the steepest slope in its partial regression plot?
```

---

# Task 2: Standardized Coefficients
```{r}
# TODO: Calculate standardized coefficients
standardized_coefs <- lm.beta(___)

# Create comparison table
comparison <- tibble(
  Predictor = names(coef(model_full))[-1],
  Raw_Coefficient = coef(model_full)[-1],
  Standardized_Beta = standardized_coefs$standardized.coefficients[-1]
) %>%
  arrange(desc(abs(Standardized_Beta)))

kable(comparison, digits = 3, 
      caption = "Raw vs. Standardized Coefficients")
```

## Visualization
```{r}
# TODO: Create bar chart of standardized coefficients
comparison %>%
  ggplot(aes(x = reorder(Predictor, Standardized_Beta),
             y = Standardized_Beta,
             fill = Standardized_Beta > 0)) +
  geom_col() +
  coord_flip() +
  labs(title = "Standardized Effect Sizes",
       x = NULL, y = "Standardized Coefficient (β*)") +
  theme(legend.position = "none")
```

**Ranking changes?**
```
TODO: Compare ranking by raw vs. standardized coefficients. 
Does the order change? Why or why not?
```

---

# Task 3: Effect Size Classification
```{r}
# TODO: Classify effect sizes
comparison <- comparison %>%
  mutate(
    Effect_Size_Category = case_when(
      abs(Standardized_Beta) < 0.3 ~ "Small",
      abs(Standardized_Beta) < 0.5 ~ "Medium",
      TRUE ~ "Large"
    )
  )

kable(comparison, digits = 3)
```

## Statistical Significance Check
```{r}
# TODO: Extract p-values and combine with effect sizes
p_values <- summary(model_full)$coefficients[-1, "Pr(>|t|)"]

comparison$P_Value <- p_values

# Flag: significant but small effect
comparison <- comparison %>%
  mutate(
    Significant = P_Value < 0.05,
    Significant_But_Small = Significant & Effect_Size_Category == "Small"
  )

kable(comparison %>% select(Predictor, Standardized_Beta, P_Value, 
                            Effect_Size_Category, Significant_But_Small),
      digits = 3)
```

**Interpretation for VP:**
```
TODO: Are any predictors statistically significant but have small 
practical effects? What should the VP do about these?
```

---

# Task 4: Prediction Scenarios
```{r}
# Define new store characteristics
new_store <- data.frame(
  foot_traffic = 500,
  employee_training_hours = 40,
  local_income = 75,
  store_age = 5,
  parking_spaces = 50
)

# TODO: Point prediction
point_pred <- predict(___, newdata = ___)

# TODO: Confidence interval (mean response)
conf_int <- predict(___, newdata = ___, 
                    interval = "confidence", level = 0.95)

# TODO: Prediction interval (individual response)
pred_int <- predict(___, newdata = ___, 
                    interval = "prediction", level = 0.95)

# Display results
cat("Point Prediction: $", round(point_pred, 1), "K\n\n", sep = "")

cat("95% Confidence Interval (mean profit):\n")
cat("  [", round(conf_int[, "lwr"], 1), ", ", 
    round(conf_int[, "upr"], 1), "]K\n\n", sep = "")

cat("95% Prediction Interval (this specific store):\n")
cat("  [", round(pred_int[, "lwr"], 1), ", ", 
    round(pred_int[, "upr"], 1), "]K\n", sep = "")
```

**Which interval to report?**
```
TODO: The VP wants to know expected profit for this new store.
Should you report the confidence or prediction interval? Why?
```

---

# Task 5: Executive Memo
```
TO: VP Martinez
FROM: [Your Name], Data Analytics
RE: Store Profit Drivers - Key Findings

VP Martinez,

Based on your request to understand profit drivers, here's what the data shows:

TODO: Write 3-4 sentences including:
1. Which 1-2 factors have the largest effects (mention standardized betas)
2. One specific, actionable recommendation
3. Uncertainty/confidence statement about predictions

[Your analysis here]
```

---

# Bonus Challenge 1: Importance Plot
```{r eval=FALSE}
# TODO: Create plot showing standardized coefficients with confidence intervals

# Extract confidence intervals for raw coefficients
conf_intervals <- confint(model_full)[-1, ]

# Convert to standardized scale (approximate)
# TODO: Calculate standardized CIs

# Plot
```

---

# Bonus Challenge 2: Impact Calculation
```{r eval=FALSE}
# TODO: Calculate profit increase from 1 SD improvement in top predictor

top_predictor <- "___"  # Fill in
top_beta_star <- ___

# 1 SD in profit units
sd_profit <- sd(store_data$profit)

# Expected increase
expected_increase <- ___

cat(sprintf("Improving %s by 1 SD increases profit by $%.1fK\n",
            top_predictor, expected_increase))
```

---

# Bonus Challenge 3: Interaction Test
```{r eval=FALSE}
# TODO: Test if training effect varies by store age

model_interaction <- lm(profit ~ foot_traffic + employee_training_hours * store_age +
                          local_income + parking_spaces,
                        data = store_data)

summary(model_interaction)

# Is the interaction significant?
# Plot the interaction if significant
```

---

# Discussion Questions

**Question 1:**
```
TODO: Why might a variable be statistically significant but practically unimportant?
```

**Question 2:**
```
TODO: When would you prefer raw coefficients over standardized?
```

**Question 3:**
```
TODO: How would you explain prediction intervals to VP Martinez?
Use plain language, no jargon.
```