---
title: "Classwork 3: Regularization Practice"
subtitle: "Overfitting Diagnosis, Ridge & LASSO Regression"
author: "Your Name Here"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

library(tidyverse)
library(car)
library(caret)
library(glmnet)
library(knitr)

theme_set(theme_minimal(base_size = 14))
```

---

# Introduction

**Scenario:** You work as a data scientist for **HomeSmart Realty**, a real estate company that wants to predict housing prices. You have data on 500 houses with 12 features including:

- Square footage
- Number of bedrooms/bathrooms
- Lot size
- Age of house
- Neighborhood characteristics
- School district quality
- Distance to amenities

Your goal is to build the best prediction model using regularization techniques.

---

# Load the Data

```{r load_data}
# Load housing data
set.seed(42)

# Generate synthetic housing data
n <- 500

housing_data <- tibble(
  house_id = 1:n,
  square_feet = rnorm(n, 2000, 500),
  bedrooms = sample(2:6, n, replace = TRUE),
  bathrooms = sample(1:4, n, replace = TRUE),
  lot_size = rnorm(n, 8000, 2000),
  age_years = sample(1:50, n, replace = TRUE),
  garage_spaces = sample(0:3, n, replace = TRUE),
  school_rating = rnorm(n, 7, 1.5),
  crime_index = rnorm(n, 50, 15),
  distance_downtown = rnorm(n, 5, 2),
  nearby_parks = sample(0:5, n, replace = TRUE),
  hoa_fees = rnorm(n, 200, 100),
  property_tax = rnorm(n, 3000, 800)
) %>%
  mutate(
    # Create price with some correlations
    price = 100000 + 
      150 * square_feet + 
      20000 * bedrooms +
      15000 * bathrooms +
      10 * lot_size +
      -2000 * age_years +
      10000 * garage_spaces +
      8000 * school_rating +
      -500 * crime_index +
      -3000 * distance_downtown +
      5000 * nearby_parks +
      rnorm(n, 0, 30000)  # Add noise
  ) %>%
  mutate(price = pmax(price, 50000))  # Minimum price

# Show first few rows
head(housing_data)
```

---

# Part 1: Overfitting Diagnosis (Exercises 1-4)

## Exercise 1: Train/Test Split

**Task:** Split the data into 80% training and 20% test sets.

```{r ex1}
# YOUR CODE HERE
# Hint: Use createDataPartition() from caret package




# Check the split
cat("Training set size:", "___", "\n")
cat("Test set size:", "___", "\n")
```

**Expected output:** Training: 400, Test: 100

---

## Exercise 2: Build Full OLS Model

**Task:** Build a linear regression model using ALL predictors (except house_id) to predict price.

```{r ex2}
# YOUR CODE HERE
# Hint: Use lm() with all predictors




# View model summary
```

**Question:** How many predictors are in your model?

---

## Exercise 3: Calculate Train and Test RMSE

**Task:** Calculate RMSE for both training and test sets. Is there overfitting?

```{r ex3}
# YOUR CODE HERE
# Calculate predictions and RMSE for both sets




cat("Training RMSE: $", "___", "\n")
cat("Test RMSE: $", "___", "\n")
cat("Degradation: ", "___", "%\n")
```

**Question:** Is the model overfitting? How can you tell?

---

## Exercise 4: Check for Multicollinearity

**Task:** Calculate VIF for all predictors. Which variables have high VIF (>5)?

```{r ex4}
# YOUR CODE HERE
# Hint: Use vif() function from car package




# Create a bar plot of VIF values
```

**Question:** Which 2-3 variables have the highest VIF? Why might they be correlated?

---

# Part 2: Ridge Regression (Exercises 5-8)

## Exercise 5: Prepare Data Matrices

**Task:** Convert the training and test data into matrix format required by glmnet.

```{r ex5}
# YOUR CODE HERE
# Create X_train, y_train, X_test, y_test
# Hint: Use model.matrix() and exclude house_id




# Check dimensions
cat("X_train dimensions:", "___", "\n")
cat("X_test dimensions:", "___", "\n")
```

---

## Exercise 6: Fit Ridge with Cross-Validation

**Task:** Use cv.glmnet() to find the optimal lambda for Ridge regression.

```{r ex6}
# YOUR CODE HERE
set.seed(123)
# Hint: alpha = 0 for Ridge




# Print optimal lambda values
cat("Lambda min:", "___", "\n")
cat("Lambda 1se:", "___", "\n")

# Plot the cross-validation results
```

---

## Exercise 7: Fit Final Ridge Model

**Task:** Fit the final Ridge model using lambda.1se and examine coefficients.

```{r ex7}
# YOUR CODE HERE
# Fit final model with lambda.1se




# Print coefficients


# Compare to OLS coefficients (optional visualization)
```

**Question:** How do Ridge coefficients compare to OLS coefficients?

---

## Exercise 8: Evaluate Ridge Performance

**Task:** Calculate training and test RMSE for Ridge. Did it improve over OLS?

```{r ex8}
# YOUR CODE HERE
# Get predictions and calculate RMSE




cat("OLS Train RMSE: $", "___", "\n")
cat("Ridge Train RMSE: $", "___", "\n")
cat("\n")
cat("OLS Test RMSE: $", "___", "\n")
cat("Ridge Test RMSE: $", "___", "\n")
cat("\n")
cat("Test improvement: $", "___", "\n")
```

**Question:** Did Ridge reduce overfitting? How can you tell?

---

# Part 3: LASSO Regression (Exercises 9-12)

## Exercise 9: Fit LASSO with Cross-Validation

**Task:** Use cv.glmnet() to find the optimal lambda for LASSO regression.

```{r ex9}
# YOUR CODE HERE
set.seed(123)
# Hint: alpha = 1 for LASSO




# Print optimal lambda values
cat("Lambda min:", "___", "\n")
cat("Lambda 1se:", "___", "\n")

# Plot the cross-validation results
# Notice the top axis shows number of features!
```

**Question:** How many features does LASSO select at lambda.1se?

---

## Exercise 10: Identify Selected Features

**Task:** Fit final LASSO model and identify which features were selected (non-zero coefficients).

```{r ex10}
# YOUR CODE HERE
# Fit final LASSO model with lambda.1se




# Extract and display selected features


# Count selected features
```

**Question:** Which features did LASSO drop? Does this make business sense?

---

## Exercise 11: Visualize Feature Selection

**Task:** Create a visualization comparing which features Ridge kept vs LASSO selected.

```{r ex11}
# YOUR CODE HERE
# Compare Ridge and LASSO coefficients side-by-side




# Create a comparison plot
```

**Question:** Which features are most important according to LASSO?

---

## Exercise 12: Final Model Comparison

**Task:** Create a comprehensive comparison table of all three methods (OLS, Ridge, LASSO).

```{r ex12}
# YOUR CODE HERE
# Calculate performance metrics for all models




# Create comparison table
comparison_table <- tibble(
  Method = c("OLS", "Ridge", "LASSO"),
  Features = c("___", "___", "___"),
  Train_RMSE = c("___", "___", "___"),
  Test_RMSE = c("___", "___", "___"),
  Overfitting = c("___", "___", "___")
)

kable(comparison_table, caption = "Model Performance Comparison")

# Visualize the comparison
```

**Question:** Which model would you recommend to HomeSmart Realty? Why?

---

# Bonus Challenges (Optional)

## Bonus 1: Coefficient Stability

**Task:** Test the stability of LASSO feature selection by fitting it on 5 different random samples of the training data.

```{r bonus1, eval=FALSE}
# YOUR CODE HERE
# Use bootstrapping or random sampling




# Which features are selected consistently?
```

---

## Bonus 2: Ridge vs LASSO Paths

**Task:** Plot the coefficient paths for both Ridge and LASSO to see how coefficients change with lambda.

```{r bonus2, eval=FALSE}
# YOUR CODE HERE
# Fit models across many lambda values
# Use plot() function




```

---

## Bonus 3: Business Recommendation

**Task:** Write a 1-paragraph recommendation to HomeSmart Realty explaining:
- Which model to use
- Why it's better than alternatives
- What the key price drivers are
- How confident you are in the predictions

```{r bonus3, eval=FALSE}
# Write your recommendation here as a comment or text

```

---

# Summary and Reflection

**Key Takeaways:**

1. **Overfitting Diagnosis:**
   - VIF detects multicollinearity
   - Train/test gap reveals overfitting
   - More predictors ≠ better predictions

2. **Ridge Regression:**
   - Shrinks all coefficients
   - Reduces overfitting
   - Keeps all features
   - Better test performance

3. **LASSO Regression:**
   - Performs feature selection
   - Sets some coefficients to zero
   - More interpretable
   - Similar performance to Ridge

4. **Practical Considerations:**
   - Always use cross-validation for λ
   - Prefer lambda.1se for stability
   - Consider business context
   - Simpler models often better

---

# Self-Check Questions

1. When should you use Ridge vs LASSO?
2. Why does LASSO set coefficients exactly to zero but Ridge doesn't?
3. What does lambda control in regularization?
4. How do you know if a model is overfitting?
5. Why use lambda.1se instead of lambda.min?

---

# Additional Resources

- **glmnet vignette:** https://glmnet.stanford.edu/articles/glmnet.html
- **Ridge vs LASSO comparison:** Statistical Learning with Applications in R
- **Cross-validation guide:** Elements of Statistical Learning

---

# Submission Instructions

**To submit your work:**

1. Complete all 12 exercises
2. Answer all questions in comments
3. Run all code chunks to verify they work
4. Knit this document to HTML
5. Submit the HTML file to Canvas

**Grading Criteria:**
- Code runs without errors (40%)
- Correct answers to exercises (40%)
- Quality of explanations (20%)

---

**Good luck! 🚀**
