---
title: "Week 1 Course Project: Advanced Regression Analysis"
subtitle: "Real Estate Price Prediction - Complete Analysis"
author: "Your Name Here"
date: "Due: End of Week 1"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

library(tidyverse)
library(car)
library(caret)
library(glmnet)
library(knitr)
library(gridExtra)

theme_set(theme_minimal(base_size = 14))
```

---

# Project Overview

## Business Context

**You are a data scientist at PremiumHomes Real Estate**, a company that operates in multiple metropolitan areas. The company wants to:

1. **Predict house prices** accurately for new listings
2. **Identify key price drivers** to advise clients
3. **Create a production model** that updates quarterly
4. **Provide insights** to the sales and marketing teams

Your task is to build and evaluate multiple regression models, ultimately recommending the best approach for production deployment.

---

## Dataset Description

You will work with a comprehensive real estate dataset containing:
- **1,000 properties** across different neighborhoods
- **20 predictor variables** (mix of numeric and categorical)
- **1 target variable:** Sale price

**Important:** This dataset includes:
- Multicollinearity (realistic correlations)
- Some irrelevant features (noise)
- Missing values (you'll handle these)
- Outliers (realistic data quality issues)

---

## Learning Objectives

By completing this project, you will demonstrate your ability to:

1. ✅ Conduct exploratory data analysis
2. ✅ Diagnose and handle overfitting
3. ✅ Detect and address multicollinearity
4. ✅ Implement Ridge, LASSO, and Elastic Net
5. ✅ Tune hyperparameters systematically
6. ✅ Compare models rigorously
7. ✅ Make data-driven recommendations
8. ✅ Communicate findings effectively

---

## Project Requirements

### Required Deliverables:

1. **Complete R Markdown file** (.Rmd) with all code and analysis
2. **Knitted HTML report** with results and visualizations
3. **Executive summary** (1-2 pages) for non-technical stakeholders
4. **Model files** saved for production deployment

### Submission Format:

- Submit as a single .zip file containing:
  - `your_name_week1_project.Rmd`
  - `your_name_week1_project.html`
  - `executive_summary.pdf` (or .docx)
  - `models/` folder with saved model objects

---

## Grading Rubric (100 points)

| Component | Points | Description |
|-----------|--------|-------------|
| **Part 1: EDA** | 15 | Data exploration, visualization, insights |
| **Part 2: Problem Diagnosis** | 15 | Overfitting, multicollinearity analysis |
| **Part 3: Ridge Regression** | 15 | Implementation, tuning, evaluation |
| **Part 4: LASSO Regression** | 15 | Implementation, feature selection |
| **Part 5: Elastic Net** | 15 | Implementation, alpha tuning |
| **Part 6: Model Comparison** | 10 | Systematic comparison, testing |
| **Part 7: Recommendation** | 10 | Business-focused recommendation |
| **Code Quality** | 5 | Clean, documented, reproducible |

**Bonus:** +10 points for exceptional work (caret integration, automation, etc.)

---

# Generate the Dataset

```{r generate_data}
# Set seed for reproducibility
set.seed(2024)

n <- 1000  # Number of properties

# Generate realistic real estate data
real_estate <- tibble(
  # Property ID
  property_id = 1:n,
  
  # Location features
  neighborhood_quality = sample(1:10, n, replace = TRUE, prob = c(rep(0.05, 5), rep(0.15, 5))),
  distance_downtown_mi = rexp(n, rate = 0.2) %>% pmin(20),
  walkability_score = rnorm(n, 65, 15) %>% pmin(100) %>% pmax(0),
  
  # Property characteristics
  square_feet = rnorm(n, 2200, 600) %>% pmax(800),
  lot_size_sqft = square_feet * runif(n, 2, 5) + rnorm(n, 0, 1000),
  bedrooms = sample(2:6, n, replace = TRUE, prob = c(0.1, 0.3, 0.3, 0.2, 0.1)),
  bathrooms = bedrooms * 0.75 + rnorm(n, 0, 0.3) %>% round() %>% pmax(1),
  garage_spaces = sample(0:3, n, replace = TRUE, prob = c(0.1, 0.3, 0.4, 0.2)),
  
  # Age and condition
  year_built = sample(1950:2023, n, replace = TRUE),
  age_years = 2024 - year_built,
  renovation_recent = rbinom(n, 1, 0.3),
  
  # Quality indicators
  school_rating = rnorm(n, 7, 1.8) %>% pmin(10) %>% pmax(1),
  crime_rate = rexp(n, 0.05) %>% pmin(100),
  air_quality_index = rnorm(n, 50, 15) %>% pmin(150) %>% pmax(0),
  
  # Amenities
  has_pool = rbinom(n, 1, 0.15),
  has_fireplace = rbinom(n, 1, 0.4),
  has_ac = rbinom(n, 1, 0.8),
  
  # Financial
  property_tax_annual = square_feet * 2.5 + rnorm(n, 0, 500),
  hoa_monthly = sample(c(0, 100, 200, 350, 500), n, replace = TRUE, 
                      prob = c(0.4, 0.2, 0.2, 0.15, 0.05)),
  
  # Market factors
  days_on_market = rpois(n, 30) %>% pmin(180),
  previous_sale_price = NA  # Noise variable (missing for all)
) %>%
  mutate(
    # Create price with realistic relationships
    price = 50000 +
      # Location
      15000 * neighborhood_quality +
      -3000 * distance_downtown_mi +
      500 * walkability_score +
      # Size
      180 * square_feet +
      8 * lot_size_sqft +
      25000 * bedrooms +
      15000 * bathrooms +
      12000 * garage_spaces +
      # Age/condition
      -800 * age_years +
      35000 * renovation_recent +
      # Quality
      12000 * school_rating +
      -400 * crime_rate +
      -200 * air_quality_index +
      # Amenities
      45000 * has_pool +
      8000 * has_fireplace +
      5000 * has_ac +
      # Random noise
      rnorm(n, 0, 40000)
  ) %>%
  mutate(
    price = pmax(price, 100000),  # Minimum price
    price = round(price, -3)  # Round to nearest $1000
  )

# Introduce some missing values (realistic scenario)
real_estate$walkability_score[sample(1:n, 30)] <- NA
real_estate$air_quality_index[sample(1:n, 25)] <- NA

# Save dataset
write_csv(real_estate, "real_estate_data.csv")

cat("Dataset generated successfully!\n")
cat("Properties:", nrow(real_estate), "\n")
cat("Features:", ncol(real_estate) - 2, "(excluding ID and price)\n")
cat("Missing values:", sum(is.na(real_estate)), "\n")
```

---

# Part 1: Exploratory Data Analysis (15 points)

## Task 1.1: Load and Inspect Data

```{r task_1_1}
# YOUR CODE HERE
# Load the dataset
# Inspect structure, dimensions, data types




# Summary statistics


# Check for missing values


# Identify numeric vs categorical variables
```

**Questions:**
1. How many observations and features do you have?
2. Which variables have missing values?
3. Which variables are categorical?

---

## Task 1.2: Target Variable Analysis

```{r task_1_2}
# YOUR CODE HERE
# Analyze the distribution of price




# Summary statistics for price


# Visualize price distribution


# Check for outliers
```

**Questions:**
1. What is the median house price?
2. Are there any extreme outliers?
3. Is the distribution approximately normal?

---

## Task 1.3: Feature Correlations

```{r task_1_3}
# YOUR CODE HERE
# Calculate correlation matrix for numeric features




# Visualize correlations (heatmap or corrplot)


# Identify highly correlated pairs (r > 0.7)


# Correlation with target variable
```

**Questions:**
1. Which features are most strongly correlated with price?
2. Which feature pairs are highly correlated with each other?
3. Does multicollinearity appear to be an issue?

---

## Task 1.4: Key Visualizations

```{r task_1_4}
# YOUR CODE HERE
# Create at least 3 meaningful visualizations:
# 1. Price vs square_feet (scatterplot)


# 2. Price by neighborhood_quality (boxplot)


# 3. Price vs age with renovation status (colored scatterplot)


# 4. Your choice - explore relationships
```

**Questions:**
1. What patterns do you observe?
2. Which features appear most important?
3. Are there any surprising relationships?

---

# Part 2: Problem Diagnosis (15 points)

## Task 2.1: Data Preparation

```{r task_2_1}
# YOUR CODE HERE
# Handle missing values




# Create train/test split (80/20)


# Prepare matrices for glmnet


# Summary of prepared data
```

**Questions:**
1. How did you handle missing values? Why?
2. What are the dimensions of train and test sets?

---

## Task 2.2: Baseline OLS Model

```{r task_2_2}
# YOUR CODE HERE
# Fit OLS regression with all features




# Training performance


# Test performance


# Calculate overfitting percentage
```

**Questions:**
1. What is the training RMSE?
2. What is the test RMSE?
3. How much overfitting is present?

---

## Task 2.3: Multicollinearity Diagnosis

```{r task_2_3}
# YOUR CODE HERE
# Calculate VIF for all predictors




# Visualize VIF values


# Identify problematic features (VIF > 5)
```

**Questions:**
1. Which features have high VIF?
2. Why are these features correlated?
3. Should you remove any features?

---

## Task 2.4: Coefficient Stability

```{r task_2_4}
# YOUR CODE HERE
# Test coefficient stability across different samples




# Compare coefficients across samples


# Visualize instability
```

**Questions:**
1. Which coefficients are most unstable?
2. Is this stability acceptable for production?

---

# Part 3: Ridge Regression (15 points)

## Task 3.1: Cross-Validation

```{r task_3_1}
# YOUR CODE HERE
# Perform cross-validation for Ridge




# Find optimal lambda


# Plot CV results
```

**Questions:**
1. What is lambda.min?
2. What is lambda.1se?
3. Which should you use for production?

---

## Task 3.2: Fit Final Ridge Model

```{r task_3_2}
# YOUR CODE HERE
# Fit Ridge with optimal lambda




# Extract coefficients


# Compare to OLS coefficients
```

**Questions:**
1. How much were coefficients shrunk?
2. Which coefficients were affected most?

---

## Task 3.3: Ridge Performance

```{r task_3_3}
# YOUR CODE HERE
# Calculate training and test RMSE




# Compare to OLS


# Visualize performance comparison
```

**Questions:**
1. Did Ridge reduce overfitting?
2. How much did test RMSE improve?
3. Is Ridge better than OLS?

---

## Task 3.4: Coefficient Path

```{r task_3_4}
# YOUR CODE HERE
# Plot Ridge coefficient path




# Mark optimal lambda
```

**Questions:**
1. Which features remain important at high lambda?
2. How do coefficients behave as lambda increases?

---

# Part 4: LASSO Regression (15 points)

## Task 4.1: LASSO Cross-Validation

```{r task_4_1}
# YOUR CODE HERE
# Perform CV for LASSO




# Find optimal lambda


# Plot CV results with feature count
```

**Questions:**
1. What is the optimal lambda?
2. How many features are selected?

---

## Task 4.2: Feature Selection

```{r task_4_2}
# YOUR CODE HERE
# Fit final LASSO model




# Identify selected features


# List dropped features
```

**Questions:**
1. Which features did LASSO select?
2. Which features were dropped?
3. Does this make business sense?

---

## Task 4.3: LASSO Performance

```{r task_4_3}
# YOUR CODE HERE
# Calculate performance metrics




# Compare to Ridge and OLS


# Create comparison table
```

**Questions:**
1. How does LASSO compare to Ridge?
2. Is the simpler model worth any performance loss?

---

## Task 4.4: Feature Importance

```{r task_4_4}
# YOUR CODE HERE
# Rank features by importance




# Visualize importance


# Identify top 5 drivers
```

**Questions:**
1. What are the top 5 price drivers?
2. How would you communicate this to stakeholders?

---

# Part 5: Elastic Net (15 points)

## Task 5.1: Tune Alpha

```{r task_5_1}
# YOUR CODE HERE
# Test multiple alpha values




# Find optimal alpha


# Plot performance vs alpha
```

**Questions:**
1. What is the optimal alpha?
2. Is it closer to Ridge or LASSO?
3. Why might this alpha be optimal?

---

## Task 5.2: Final Elastic Net Model

```{r task_5_2}
# YOUR CODE HERE
# Fit Elastic Net with optimal alpha




# Extract coefficients


# Count selected features
```

**Questions:**
1. How many features did Elastic Net select?
2. Are they the same as LASSO?

---

## Task 5.3: Stability Analysis

```{r task_5_3}
# YOUR CODE HERE
# Test Elastic Net stability




# Compare to LASSO stability


# Visualize
```

**Questions:**
1. Is Elastic Net more stable than LASSO?
2. Which features are consistently selected?

---

## Task 5.4: Performance Evaluation

```{r task_5_4}
# YOUR CODE HERE
# Calculate all performance metrics




# Create comprehensive comparison
```

**Questions:**
1. How does Elastic Net rank overall?
2. Is it worth the added complexity?

---

# Part 6: Model Comparison (10 points)

## Task 6.1: Comprehensive Comparison

```{r task_6_1}
# YOUR CODE HERE
# Create detailed comparison table




# Include:
# - Method name
# - Hyperparameters
# - Features selected
# - Training RMSE
# - Test RMSE
# - R-squared
# - Interpretability score
```

**Questions:**
1. Which model performs best on test data?
2. Which model is most interpretable?
3. What trade-offs exist?

---

## Task 6.2: Visualization

```{r task_6_2}
# YOUR CODE HERE
# Create comprehensive comparison visualizations
# 1. RMSE comparison (bar chart)


# 2. Feature selection comparison (venn diagram or table)


# 3. Prediction accuracy (actual vs predicted scatter)


# 4. Your choice
```

**Questions:**
1. Which visualization is most informative?
2. What story do the visualizations tell?

---

## Task 6.3: Statistical Testing

```{r task_6_3}
# YOUR CODE HERE
# Perform statistical tests if appropriate




# Are performance differences significant?
```

**Questions:**
1. Are the differences statistically significant?
2. Or are they within noise?

---

# Part 7: Business Recommendation (10 points)

## Task 7.1: Model Selection

```{r task_7_1}
# YOUR CODE HERE
# Save your recommended model




# Document hyperparameters


# Save feature importance
```

**Your Recommendation:**

Write your recommendation here (3-5 paragraphs):

```
RECOMMENDED MODEL: [Your choice]

RATIONALE:
1. Performance: [Explain test RMSE, comparison to alternatives]

2. Interpretability: [Discuss feature selection, stakeholder communication]

3. Stability: [Discuss coefficient stability, production reliability]

4. Business Value: [Explain practical benefits to PremiumHomes]

5. Implementation: [Discuss deployment considerations]

KEY FEATURES IDENTIFIED:
- [Feature 1]: [Impact on price]
- [Feature 2]: [Impact on price]
- [Feature 3]: [Impact on price]
...

EXPECTED ACCURACY:
- Average prediction error: $X,XXX
- Confidence interval: $X,XXX to $X,XXX

RECOMMENDATIONS FOR USE:
- [How to use model]
- [What to watch out for]
- [When to retrain]
```

---

## Task 7.2: Stakeholder Communication

Create visualizations and talking points for three audiences:

### For Executives:

```{r task_7_2_exec}
# YOUR CODE HERE
# Create 1-2 executive-level visualizations




# Key metrics dashboard
```

**Executive Summary (3 bullet points):**
- 
- 
- 

---

### For Sales Team:

```{r task_7_2_sales}
# YOUR CODE HERE
# Create practical guidance for pricing




# Feature importance for sales
```

**Sales Team Guidance:**
- 
- 
- 

---

### For Technical Team:

```{r task_7_2_tech}
# YOUR CODE HERE
# Technical specifications




# Deployment requirements
```

**Technical Documentation:**
- Model: 
- Hyperparameters: 
- Dependencies: 
- Retraining schedule: 

---

# Bonus Challenges (Optional, +10 points)

## Bonus 1: caret Integration

```{r bonus_1, eval=FALSE}
# YOUR CODE HERE
# Implement using caret package




# Automated hyperparameter tuning


# Model comparison using resamples
```

**Benefit of caret approach:**

---

## Bonus 2: Advanced Diagnostics

```{r bonus_2, eval=FALSE}
# YOUR CODE HERE
# Residual analysis




# Influential points


# Model assumptions
```

**Key findings:**

---

## Bonus 3: Automated Pipeline

```{r bonus_3, eval=FALSE}
# YOUR CODE HERE
# Create automated model selection function




# Production deployment script
```

**Documentation:**

---

# Final Checklist

Before submission, verify:

- [ ] All code chunks run without errors
- [ ] All questions are answered
- [ ] Visualizations are clear and labeled
- [ ] Business recommendation is complete
- [ ] Code is well-commented
- [ ] Document knits to HTML successfully
- [ ] Executive summary created
- [ ] Model files saved
- [ ] All files zipped together

---

# Submission

**Submit to:** [Course Platform]

**Deadline:** End of Week 1

**File name format:** `LastName_FirstName_Week1_Project.zip`

**Zip contents:**
1. `LastName_FirstName_Week1_Project.Rmd`
2. `LastName_FirstName_Week1_Project.html`
3. `executive_summary.pdf` (or .docx)
4. `models/` folder with saved models
5. `real_estate_data.csv` (the generated dataset)

---

# Evaluation Criteria

Your project will be evaluated on:

1. **Technical Correctness** (40%)
   - Proper implementation of methods
   - Correct use of validation techniques
   - Appropriate hyperparameter tuning

2. **Analysis Quality** (30%)
   - Depth of exploration
   - Quality of visualizations
   - Insightful interpretations

3. **Communication** (20%)
   - Clear explanations
   - Professional presentation
   - Effective stakeholder communication

4. **Code Quality** (10%)
   - Clean, readable code
   - Good documentation
   - Reproducibility

---

# Tips for Success

1. **Start early** - This is a comprehensive project
2. **Be systematic** - Follow the parts in order
3. **Document everything** - Future you will thank present you
4. **Think like a data scientist** - Not just running code, but understanding why
5. **Communicate clearly** - Technical skills + communication = success
6. **Ask questions** - Use office hours if stuck
7. **Be creative** - Show your unique insights in bonus sections

---

**Good luck! 🚀**

**Remember:** This project demonstrates real-world data science skills. Take pride in your work!
