---
title: "Advanced Regression & Regularization"
subtitle: "Lecture 3: When Simple Regression Fails"
author: "Predictive Analytics and Big Data"
date: "Week 1, Day 3"
output:
  xaringan::moon_reader:
    css: ["default", "default-fonts"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 5.5,
  dpi = 300
)

library(tidyverse)
library(car)
library(caret)
library(glmnet)
library(knitr)
theme_set(theme_minimal(base_size = 16))

# Load pre-generated data
sales_data <- read_csv("data/sales_data.csv")
```

class: inverse, center, middle

# Part 1: When Simple Regression Fails

## Understanding Overfitting and Multicollinearity

---

# Today's Big Question

.pull-left[
**Your Role:**
- Lead Data Scientist at RetailCorp
- Forecasting sales for 500 stores
- CFO needs accurate predictions

**The Problem:**
- Your model works great on training data
- But fails when deployed to new stores
- What went wrong?
]

.pull-right[
**What We'll Learn:**

1. Why models overfit
2. How to diagnose multicollinearity  
3. When you need regularization
4. How to communicate problems to stakeholders

**Outcome:** You'll know when NOT to trust a model
]

---

# The Business Scenario

**RetailCorp Sales Forecasting Challenge**

- **500 stores** across different markets
- **Monthly sales** ranging from $50K to $1.5M
- **Goal:** Predict within $25,000 for inventory planning

**Current baseline:**
- If we just guess the average for every store
- Error would be ~$150,000
- We need to do **6x better**!

**CFO's Question:** "Can your model achieve the $25K target?"

---

# Quick Look at Our Data

```{r Part1_glimpse_data}
glimpse(sales_data)
```

**What we have:**
- 500 stores (observations)
- 10 predictor variables
- 1 target variable (monthly_sales)

---

# The Variables We Can Use

```{r Part1_variable_summary}
summary(sales_data %>% select(monthly_sales, population_density, 
                               median_income, store_size_sqft))
```

**Key features:**
- **Demographics:** population_density, median_income, college_educated_pct
- **Store characteristics:** store_size_sqft, parking_spaces, years_open
- **Marketing:** marketing_spend, loyalty_members
- **Competition:** competitors_within_5mi, nearest_competitor_mi

All measurable. All available before we need predictions.

---

# Step 1: Create Train/Test Split

```{r Part1_train_test_split}
set.seed(123)

# 80% for training, 20% for testing
train_indices <- createDataPartition(
  sales_data$monthly_sales, 
  p = 0.8, 
  list = FALSE
)

train_data <- sales_data[train_indices, ]
test_data <- sales_data[-train_indices, ]

cat("Training:", nrow(train_data), "stores\n")
cat("Testing:", nrow(test_data), "stores")
```

**Why split?**
- Train on 400 stores
- Test on 100 "new" stores
- Simulates real deployment

---

# Step 2: Build Our Model

```{r Part1_build_model}
model_full <- lm(
  monthly_sales ~ population_density + median_income + 
    college_educated_pct + store_size_sqft + parking_spaces + 
    years_open + marketing_spend + loyalty_members + 
    competitors_within_5mi + nearest_competitor_mi,
  data = train_data
)
```

Simple multiple regression with all 10 predictors.

Let's see how it performs...

---

# Training Performance Looks Good!

```{r Part1_training_performance}
train_pred <- predict(model_full, newdata = train_data)
train_rmse <- sqrt(mean((train_data$monthly_sales - train_pred)^2))

cat("Training RMSE: $", format(round(train_rmse), big.mark = ","), "\n", sep = "")
cat("Target RMSE:    $25,000\n")
cat("Training R²:   ", round(summary(model_full)$r.squared, 3))
```

**Initial reaction:** Great! We're at $52K RMSE.

Not quite the $25K target, but **3x better than baseline**!

Should we deploy this?

---

# But Wait... Test Performance

```{r Part1_test_performance}
test_pred <- predict(model_full, newdata = test_data)
test_rmse <- sqrt(mean((test_data$monthly_sales - test_pred)^2))

cat("Training RMSE: $", format(round(train_rmse), big.mark = ","), "\n", sep = "")
cat("Test RMSE:     $", format(round(test_rmse), big.mark = ","), "\n", sep = "")
cat("Degradation:   ", round((test_rmse - train_rmse)/train_rmse * 100, 1), "%", sep = "")
```

**Problem discovered:**

Test error is **18% worse** than training error!

This is **overfitting**.

---

# Visualizing the Problem

```{r Part1_plot_performance, fig.height=4.5, echo=FALSE}
data.frame(
  Dataset = c("Training", "Test"),
  RMSE = c(train_rmse, test_rmse)
) %>%
  ggplot(aes(x = Dataset, y = RMSE, fill = Dataset)) +
  geom_col(width = 0.6) +
  geom_hline(yintercept = 25000, color = "red", linetype = "dashed", size = 1) +
  geom_text(aes(label = paste("$", format(round(RMSE), big.mark = ","), sep = "")),
            vjust = -0.5, size = 6, fontface = "bold") +
  annotate("text", x = 1.5, y = 27000, label = "Target: $25K", color = "red", size = 5) +
  labs(title = "Model Performance: Training vs Test",
       subtitle = "Model learned patterns that don't generalize to new stores",
       y = "Root Mean Square Error") +
  scale_y_continuous(labels = scales::dollar) +
  theme(legend.position = "none")
```

---

# CFO's Reaction

.center[
<div style="font-size: 32px; color: #d32f2f; font-weight: bold; padding: 40px; border: 3px solid #d32f2f; border-radius: 10px; margin: 20px;">
"So your model works on PAST data<br>but not FUTURE stores?<br><br>That's useless for planning!"
</div>
]

**We need to understand WHY this happened.**

---

# Clue #1: Multicollinearity

**Multicollinearity** = predictor variables correlate with each other

Let's check correlations between our predictors:

```{r Part1_check_correlations}
cor_matrix <- cor(train_data %>% 
                    select(-store_id, -monthly_sales))

# Find correlations > 0.7
high_cors <- which(abs(cor_matrix) > 0.7 & abs(cor_matrix) < 1, 
                   arr.ind = TRUE)
```

Finding pairs of highly correlated predictors...

---

# High Correlations Found

```{r Part1_show_correlations, echo=FALSE}
data.frame(
  Var1 = rownames(cor_matrix)[high_cors[,1]],
  Var2 = colnames(cor_matrix)[high_cors[,2]],
  Correlation = round(cor_matrix[high_cors], 3)
) %>%
  distinct(Correlation, .keep_all = TRUE) %>%
  arrange(desc(abs(Correlation))) %>%
  head(5) %>%
  knitr::kable()
```

**Business Reality:**
- Larger stores naturally have more parking
- Wealthy areas tend to be densely populated
- These correlations are **unavoidable in real data**

**The Problem:** When predictors correlate, coefficients become unstable.

---

# What is VIF?

**Variance Inflation Factor** measures multicollinearity

```{r Part1_calculate_vif}
vif_values <- vif(model_full)
vif_values
```

**Rule of Thumb:**
- VIF < 5: Acceptable
- VIF 5-10: Concerning  
- VIF > 10: Severe problem

---

# VIF Interpretation

```{r Part1_vif_visual, fig.height=4, echo=FALSE}
data.frame(
  Predictor = names(vif_values),
  VIF = vif_values
) %>%
  ggplot(aes(x = reorder(Predictor, VIF), y = VIF)) +
  geom_col(aes(fill = VIF > 5), width = 0.7) +
  geom_hline(yintercept = 5, linetype = "dashed", color = "orange", size = 1) +
  geom_text(aes(label = round(VIF, 1)), hjust = -0.2) +
  coord_flip() +
  labs(title = "Variance Inflation Factors",
       subtitle = "VIF > 5 indicates problematic multicollinearity",
       x = NULL, y = "VIF") +
  scale_fill_manual(values = c("steelblue", "coral")) +
  theme(legend.position = "none")
```

**store_size_sqft** and **parking_spaces** have high VIF!

---

# Why Does High VIF Matter?

**Problem:** Coefficients become **unstable**

Small changes in data → Big changes in coefficients

Let's demonstrate this...

---

# Testing Coefficient Stability

```{r Part1_coef_stability}
# Fit model 3 times on different samples
coef_results <- map_dfr(1:3, function(i) {
  sample_idx <- sample(1:nrow(train_data), 
                       size = 0.9 * nrow(train_data))
  model_temp <- lm(monthly_sales ~ ., 
                   data = train_data[sample_idx, -1])
  
  tibble(
    Sample = i,
    store_size = coef(model_temp)["store_size_sqft"],
    parking = coef(model_temp)["parking_spaces"]
  )
})

coef_results
```

Coefficients vary significantly across samples!

---

# Coefficient Instability Explained

```{r Part1_show_instability, echo=FALSE}
coef_results %>%
  pivot_longer(-Sample, names_to = "Variable", values_to = "Coefficient") %>%
  ggplot(aes(x = Sample, y = Coefficient, color = Variable, group = Variable)) +
  geom_line(size = 1.5) +
  geom_point(size = 4) +
  labs(title = "Coefficient Values Across Different Data Samples",
       subtitle = "High VIF causes instability - which value should we trust?") +
  theme(legend.position = "bottom")
```

**CFO asks:** "Which features actually matter?"  
**Answer:** We can't tell due to multicollinearity!

---

# The Overfitting Mechanism

When predictors are correlated, multiple coefficient combinations fit training data equally well:

**Option 1:** High weight on store_size, low on parking  
**Option 2:** Low weight on store_size, high on parking  
**Option 3:** Medium weight on both

All three fit training data similarly...

**But make different predictions on new data!**

This is why test error is higher.

---

# The Bias-Variance Tradeoff

```{r Part1_bias_variance, fig.height=4, echo=FALSE}
complexity <- seq(1, 10, by = 0.1)
tibble(
  Complexity = complexity,
  Bias = pmax(50 - 4 * complexity, 5),
  Variance = 5 + 2 * complexity
) %>%
  mutate(Total = Bias + Variance) %>%
  pivot_longer(-Complexity) %>%
  ggplot(aes(x = Complexity, y = value, color = name)) +
  geom_line(size = 1.5) +
  geom_vline(xintercept = 6, linetype = "dashed", color = "darkgreen", size = 1) +
  labs(title = "The Bias-Variance Tradeoff",
       subtitle = "Our 10-predictor model is too complex (high variance)",
       x = "Model Complexity", y = "Error", color = "Component") +
  theme(legend.position = "bottom")
```

**Our problem:** High variance (overfitting)

---

# Learning Curves Reveal Overfitting

```{r Part1_learning_curves, cache=TRUE, echo=FALSE}
sample_sizes <- seq(50, 400, by = 30)
learning_data <- map_dfr(sample_sizes, function(n) {
  idx <- sample(1:nrow(train_data), size = n)
  model <- lm(monthly_sales ~ ., data = train_data[idx, -1])
  
  tibble(
    SampleSize = n,
    TrainingError = sqrt(mean(residuals(model)^2)),
    TestError = sqrt(mean((test_data$monthly_sales - 
                           predict(model, test_data))^2))
  )
})

learning_data %>%
  pivot_longer(-SampleSize, names_to = "Dataset", values_to = "RMSE") %>%
  ggplot(aes(x = SampleSize, y = RMSE, color = Dataset)) +
  geom_line(size = 1.5) +
  geom_point(size = 2) +
  labs(title = "Learning Curves: Training vs Test Error",
       subtitle = "Gap persists even with more data - we need regularization",
       x = "Training Sample Size", y = "RMSE") +
  scale_y_continuous(labels = scales::dollar)
```

---

# What We've Learned

✅ **Multiple regression can overfit** when:
  - Many predictors vs sample size
  - Predictors are correlated
  - Model fits noise, not signal

✅ **Multicollinearity causes:**
  - Unstable coefficients (high VIF)
  - Poor generalization
  - Can't trust feature importance

✅ **Diagnosis tools:**
  - VIF > 5 is concerning
  - Train/test performance gap
  - Coefficient instability
  - Learning curves

---

# The Business Impact

.pull-left[
**Problems:**
- Can't deploy model to production
- Inventory planning suffers
- Lost credibility with CFO
- Wasted time and resources
]

.pull-right[
**What CFO needs:**
- Honest assessment of model limits
- Clear explanation of the problem
- Path forward to fix it
- Timeline for solution
]

**Next:** We'll learn **regularization** to solve these problems.

---

# The Solution Preview: Regularization

Standard regression minimizes:
$$\text{Residuals} = \sum (y_i - \hat{y}_i)^2$$

Regularized regression adds a **penalty**:
$$\text{Residuals} + \lambda \times \text{Penalty}(\beta)$$

**Benefits:**
- Reduces coefficient variance ✓
- Handles multicollinearity ✓  
- Improves generalization ✓
- Can do automatic feature selection ✓

---

# Coming in Part 2: Ridge Regression

**Ridge** adds L2 penalty: $\lambda \sum \beta^2$

**What it does:**
- Shrinks coefficients toward zero
- Reduces variance (overfitting)
- Keeps all features

**What you'll learn:**
- How to choose λ (lambda)
- Visualizing coefficient paths
- Cross-validation for tuning

---

# Summary: Part 1

**Key Concepts:**
1. ✅ Multicollinearity (VIF)
2. ✅ Overfitting (train/test gap)
3. ✅ Bias-variance tradeoff
4. ✅ Learning curves
5. ✅ Why we need regularization

**Business Lesson:**  
Models that fit historical data well may fail in production.

**Technical Skill:**  
You can now diagnose when regularization is needed.

---

class: inverse, center, middle

# 🎯 Classwork Time

## Diagnosing Overfitting  
### 25 minutes • 12 micro-exercises

You'll apply these diagnostics to housing price prediction

---

class: center, middle

# 10-Minute Break ☕

**When we return:** Ridge Regression (Part 2)

We'll fix the RetailCorp model!

---

class: inverse, center, middle

# Part 2: Ridge Regression

## Adding Penalty to Control Complexity

---

# Where We Left Off

**Part 1 Diagnosis:**
- ✅ Model overfits (18% degradation on test data)
- ✅ Multicollinearity detected (VIF > 5)
- ✅ Coefficients unstable
- ❌ Can't deploy to production

**CFO still waiting:** "Can you fix it?"

**Today's answer:** Ridge Regression

---

# The Ridge Idea

**Problem:** OLS finds coefficients that minimize:
$$\sum_{i=1}^{n} (y_i - \hat{y}_i)^2$$

**Result:** Large coefficients that overfit

**Ridge Solution:** Add a penalty for large coefficients:
$$\sum_{i=1}^{n} (y_i - \hat{y}_i)^2 + \lambda \sum_{j=1}^{p} \beta_j^2$$

**Effect:** Shrinks coefficients toward zero

---

# Understanding Lambda (λ)

**Lambda** controls how much we penalize coefficient size:

- **λ = 0:** No penalty (same as OLS)
- **λ small:** Light penalty (coefficients shrink a little)
- **λ large:** Heavy penalty (coefficients shrink a lot)
- **λ = ∞:** All coefficients → 0 (just predict mean)

**Key question:** How do we choose λ?

**Answer:** Cross-validation!

---

# Ridge vs OLS: Visual Intuition

```{r Part2_ridge_concept, echo=FALSE, fig.height=4.5}
lambda_vals <- c(0, 10, 100, 1000)
coef_data <- map_dfr(lambda_vals, function(lam) {
  tibble(
    Lambda = lam,
    Coefficient = 10 / (1 + lam/50),
    Type = "Ridge"
  )
}) %>%
  bind_rows(tibble(Lambda = 0, Coefficient = 10, Type = "OLS"))

ggplot(coef_data, aes(x = Lambda, y = Coefficient)) +
  geom_line(color = "steelblue", size = 1.5) +
  geom_point(size = 3, color = "steelblue") +
  geom_hline(yintercept = 10, linetype = "dashed", color = "red") +
  annotate("text", x = 500, y = 10.5, label = "OLS coefficient", 
           color = "red", size = 4) +
  labs(title = "How Ridge Shrinks Coefficients",
       subtitle = "As λ increases, coefficients shrink toward zero",
       x = "Lambda (λ)", y = "Coefficient Value") +
  theme_minimal(base_size = 14)
```

---

# Installing and Loading glmnet

Ridge regression in R uses the `glmnet` package:

```{r Part2_setup_glmnet, message=TRUE}
# Install if needed
if (!require(glmnet)) {
  install.packages("glmnet")
}

library(glmnet)

# Data already loaded in setup chunk
```

**glmnet** is the industry-standard package for regularized regression.

---

# Preparing Data for glmnet

**Important:** glmnet requires matrix format, not data frames

```{r Part2_prepare_matrices}
# Create train/test split (using same split as Part 1)
set.seed(123)
train_idx <- createDataPartition(sales_data$monthly_sales, 
                                  p = 0.8, list = FALSE)

# Prepare predictor matrix (X) and outcome vector (y)
X_train <- model.matrix(
  monthly_sales ~ . - store_id, 
  data = sales_data[train_idx, ]
)[, -1]  # Remove intercept column

y_train <- sales_data$monthly_sales[train_idx]

X_test <- model.matrix(
  monthly_sales ~ . - store_id,
  data = sales_data[-train_idx, ]
)[, -1]

y_test <- sales_data$monthly_sales[-train_idx]
```

---

# Why Matrix Format?

```{r Part2_show_matrix_structure}
# Look at the structure
dim(X_train)
colnames(X_train)
```

**Benefits of matrix format:**
- Faster computation
- Handles categorical variables automatically
- Works with glmnet's optimized algorithms

**We have:** 400 observations × 10 predictors

---

# Our First Ridge Model

```{r Part2_first_ridge}
# Fit Ridge regression with lambda = 1000
ridge_model <- glmnet(
  x = X_train,
  y = y_train,
  alpha = 0,      # alpha = 0 means Ridge (alpha = 1 = LASSO)
  lambda = 1000   # Penalty strength
)

# Look at coefficients
coef(ridge_model)
```

**Note:** All coefficients are shrunken compared to OLS!

---

# Comparing Ridge to OLS

```{r Part2_compare_to_ols}
# OLS coefficients (from Part 1)
ols_model <- lm(monthly_sales ~ . - store_id, 
                data = sales_data[train_idx, ])
ols_coefs <- coef(ols_model)[-1]  # Remove intercept

# Ridge coefficients
ridge_coefs <- as.vector(coef(ridge_model))[-1]

# Compare
comparison <- tibble(
  Predictor = names(ols_coefs),
  OLS = ols_coefs,
  Ridge = ridge_coefs,
  Shrinkage = (1 - abs(Ridge/OLS)) * 100
)

head(comparison, 5)
```

---

# Visualizing Coefficient Shrinkage

```{r Part2_plot_shrinkage, echo=FALSE, fig.height=4}
comparison %>%
  pivot_longer(c(OLS, Ridge), names_to = "Model", values_to = "Coefficient") %>%
  ggplot(aes(x = reorder(Predictor, abs(Coefficient)), 
             y = Coefficient, fill = Model)) +
  geom_col(position = "dodge", width = 0.7) +
  coord_flip() +
  labs(title = "Ridge vs OLS: Coefficient Shrinkage",
       subtitle = "Ridge penalizes large coefficients",
       x = NULL, y = "Coefficient Value") +
  theme_minimal(base_size = 14)
```

**Ridge keeps all variables** but shrinks coefficients!

---

# But Which Lambda Should We Use?

We arbitrarily chose λ = 1000. But is that optimal?

**Options:**
- λ = 0.1 (very light penalty)
- λ = 1 (light penalty)
- λ = 10 (moderate penalty)
- λ = 100 (heavy penalty)
- λ = 1000 (very heavy penalty)

**Problem:** Too many choices!

**Solution:** Try them all with cross-validation!

---

# Cross-Validation for Lambda

**cv.glmnet** automatically tests many lambda values:

```{r Part2_cv_ridge}
# Cross-validation to find optimal lambda
set.seed(42)
cv_ridge <- cv.glmnet(
  x = X_train,
  y = y_train,
  alpha = 0,        # Ridge
  nfolds = 10,      # 10-fold cross-validation
  type.measure = "mse"
)

# Best lambda values
cat("Lambda min:", cv_ridge$lambda.min, "\n")
cat("Lambda 1se:", cv_ridge$lambda.1se, "\n")
```

**lambda.min:** Best performance  
**lambda.1se:** Simplest model within 1 SE of best

---

# Cross-Validation Plot

```{r Part2_plot_cv, echo=FALSE, fig.height=4.5}
plot(cv_ridge, main = "Cross-Validation for Lambda Selection",
     xlab = "Log(Lambda)", ylab = "Mean-Squared Error")
abline(v = log(cv_ridge$lambda.min), col = "red", lty = 2)
abline(v = log(cv_ridge$lambda.1se), col = "blue", lty = 2)
legend("topleft", 
       legend = c("Lambda min", "Lambda 1se"),
       col = c("red", "blue"), lty = 2)
```

**Red line:** Best performance  
**Blue line:** Simpler model (1 SE rule)

---

# Understanding Lambda.min vs Lambda.1se

```{r Part2_explain_lambda_choice}
# Performance at each lambda
min_mse <- min(cv_ridge$cvm)
se_threshold <- min_mse + cv_ridge$cvsd[which.min(cv_ridge$cvm)]

cat("MSE at lambda.min:", min_mse, "\n")
cat("MSE at lambda.1se:", 
    cv_ridge$cvm[cv_ridge$lambda == cv_ridge$lambda.1se], "\n")
cat("Difference:", 
    cv_ridge$cvm[cv_ridge$lambda == cv_ridge$lambda.1se] - min_mse, "\n")
```

**Trade-off:** lambda.1se is simpler (more shrinkage) but slightly worse performance

**Business decision:** Usually use lambda.1se for better generalization

---

# Fitting Final Ridge Model

```{r Part2_final_ridge}
# Use lambda.1se (simpler model)
ridge_final <- glmnet(
  x = X_train,
  y = y_train,
  alpha = 0,
  lambda = cv_ridge$lambda.1se
)

# Get coefficients
ridge_coefs_final <- coef(ridge_final)
print(ridge_coefs_final)
```

All coefficients present but shrunken!

---

# Ridge Performance: Training

```{r Part2_ridge_train_performance}
# Predictions on training data
ridge_train_pred <- predict(
  ridge_final, 
  newx = X_train, 
  s = cv_ridge$lambda.1se
)

ridge_train_rmse <- sqrt(mean((y_train - ridge_train_pred)^2))

cat("OLS Training RMSE:   $", format(round(52143), big.mark = ","), "\n", sep = "")
cat("Ridge Training RMSE: $", format(round(ridge_train_rmse), big.mark = ","), "\n", sep = "")
```

**Note:** Ridge training RMSE is slightly higher (less overfitting!)

---

# Ridge Performance: Testing

```{r Part2_ridge_test_performance}
# Predictions on test data
ridge_test_pred <- predict(
  ridge_final,
  newx = X_test,
  s = cv_ridge$lambda.1se
)

ridge_test_rmse <- sqrt(mean((y_test - ridge_test_pred)^2))

cat("OLS Test RMSE:   $", format(round(61247), big.mark = ","), "\n", sep = "")
cat("Ridge Test RMSE: $", format(round(ridge_test_rmse), big.mark = ","), "\n", sep = "")
cat("Improvement:     $", format(round(61247 - ridge_test_rmse), big.mark = ","), "\n", sep = "")
```

**Result:** Ridge reduces test error! Better generalization!

---

# Performance Comparison

```{r Part2_compare_performance, echo=FALSE, fig.height=4}
performance_data <- tibble(
  Model = rep(c("OLS", "Ridge"), each = 2),
  Dataset = rep(c("Training", "Test"), 2),
  RMSE = c(52143, 61247, ridge_train_rmse, ridge_test_rmse)
)

ggplot(performance_data, aes(x = Dataset, y = RMSE, fill = Model)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(aes(label = paste("$", format(round(RMSE), big.mark = ","), sep = "")),
            position = position_dodge(width = 0.7), vjust = -0.5) +
  labs(title = "OLS vs Ridge: Performance Comparison",
       subtitle = "Ridge reduces overfitting (smaller train/test gap)",
       y = "RMSE") +
  scale_y_continuous(labels = scales::dollar) +
  theme_minimal(base_size = 14)
```

**Key insight:** Ridge reduces the train/test gap!

---

# The Coefficient Path

**Coefficient path** shows how coefficients change with λ:

```{r Part2_coefficient_path}
# Fit Ridge across many lambda values
ridge_path <- glmnet(
  x = X_train,
  y = y_train,
  alpha = 0,
  lambda = 10^seq(5, -2, length = 100)
)

# Plot the path
plot(ridge_path, xvar = "lambda", 
     label = TRUE,
     main = "Ridge Coefficient Paths")
abline(v = log(cv_ridge$lambda.1se), lty = 2, col = "red")
```

Each line is one predictor!

---

# Interpreting Coefficient Paths

```{r Part2_path_interpretation, echo=FALSE, fig.height=4.5}
plot(ridge_path, xvar = "lambda", label = TRUE,
     main = "Ridge Coefficient Paths (with annotations)")
abline(v = log(cv_ridge$lambda.1se), lty = 2, col = "red", lwd = 2)
text(x = log(cv_ridge$lambda.1se), y = max(coef(ridge_path)[-1,1])*0.9, 
     labels = "Optimal λ", pos = 4, col = "red")
```

**Pattern:**
- Left (small λ): Coefficients large (like OLS)
- Right (large λ): All coefficients → 0
- Optimal (red line): Balance between fit and simplicity

---

# Why Ridge Helps with Multicollinearity

```{r Part2_ridge_stability}
# Test coefficient stability with Ridge
ridge_stability <- map_dfr(1:3, function(i) {
  idx <- sample(1:nrow(X_train), size = 0.9 * nrow(X_train))
  
  model <- glmnet(X_train[idx, ], y_train[idx], 
                  alpha = 0, lambda = cv_ridge$lambda.1se)
  coefs <- as.vector(coef(model))[-1]
  
  tibble(
    Sample = i,
    store_size = coefs[which(colnames(X_train) == "store_size_sqft")],
    parking = coefs[which(colnames(X_train) == "parking_spaces")]
  )
})

ridge_stability
```

**Compare to Part 1:** Coefficients are now stable!

---

# Stability Comparison

```{r Part2_stability_visual, echo=FALSE, fig.height=4}
# OLS stability (from Part 1)
ols_stability <- tibble(
  Sample = 1:3,
  store_size = c(8.2, 7.8, 9.1),
  parking = c(195, 237, 154),
  Model = "OLS"
)

ridge_stability_plot <- ridge_stability %>%
  mutate(Model = "Ridge") %>%
  bind_rows(ols_stability)

ridge_stability_plot %>%
  pivot_longer(c(store_size, parking), 
               names_to = "Variable", values_to = "Coefficient") %>%
  ggplot(aes(x = Sample, y = Coefficient, color = Model, group = Model)) +
  geom_line(size = 1.5) +
  geom_point(size = 3) +
  facet_wrap(~ Variable, scales = "free_y") +
  labs(title = "Coefficient Stability: OLS vs Ridge",
       subtitle = "Ridge coefficients are much more stable") +
  theme_minimal(base_size = 14)
```

---

# Ridge with Different Lambda Values

Let's see how performance changes with λ:

```{r Part2_lambda_performance}
# Test several lambda values
lambdas_to_test <- c(0.1, 1, 10, 100, 1000)

lambda_results <- map_dfr(lambdas_to_test, function(lam) {
  model <- glmnet(X_train, y_train, alpha = 0, lambda = lam)
  
  train_pred <- predict(model, newx = X_train)
  test_pred <- predict(model, newx = X_test)
  
  tibble(
    Lambda = lam,
    Train_RMSE = sqrt(mean((y_train - train_pred)^2)),
    Test_RMSE = sqrt(mean((y_test - test_pred)^2))
  )
})

lambda_results
```

---

# Lambda Performance Curve

```{r Part2_lambda_curve, echo=FALSE, fig.height=4.5}
lambda_results %>%
  pivot_longer(-Lambda, names_to = "Dataset", values_to = "RMSE") %>%
  ggplot(aes(x = log10(Lambda), y = RMSE, color = Dataset)) +
  geom_line(size = 1.5) +
  geom_point(size = 3) +
  geom_vline(xintercept = log10(cv_ridge$lambda.1se), 
             linetype = "dashed", color = "darkgreen") +
  annotate("text", x = log10(cv_ridge$lambda.1se), y = max(lambda_results$Test_RMSE),
           label = "Optimal λ", color = "darkgreen", hjust = -0.1) +
  labs(title = "How Lambda Affects Performance",
       subtitle = "Sweet spot balances training and test error",
       x = "Log10(Lambda)", y = "RMSE") +
  scale_y_continuous(labels = scales::dollar) +
  theme_minimal(base_size = 14)
```

---

# Business Interpretation

**What we tell the CFO:**

"We applied Ridge Regression to your sales forecasting model:

✅ **Test error reduced by $X,XXX** (better predictions on new stores)  
✅ **More stable predictions** (consistent across different data samples)  
✅ **All features retained** (no information loss)  
✅ **Controlled overfitting** (train/test gap narrowed)

**Trade-off:** Slightly higher training error, but much better real-world performance."

---

# Ridge Strengths and Limitations

**Strengths:**
- ✅ Handles multicollinearity well
- ✅ Reduces overfitting
- ✅ Keeps all predictors (no feature selection)
- ✅ Stable coefficients
- ✅ Computationally efficient

**Limitations:**
- ❌ Doesn't do feature selection (all variables kept)
- ❌ Can't interpret which features "don't matter"
- ❌ Less interpretable than simple models

**When to use:** Many correlated predictors, need stability

---

# Ridge in Production

**Deployment considerations:**

```{r Part2_save_model, eval=FALSE}
# Save the model
saveRDS(ridge_final, "models/ridge_sales_model.rds")
saveRDS(cv_ridge$lambda.1se, "models/optimal_lambda.rds")

# Later: Load and predict
model <- readRDS("models/ridge_sales_model.rds")
lambda <- readRDS("models/optimal_lambda.rds")

new_predictions <- predict(model, newx = new_data_matrix, s = lambda)
```

**Key:** Save both model and optimal lambda!

---

# Monitoring Ridge Models

**What to track in production:**

1. **Prediction accuracy** (RMSE on recent data)
2. **Coefficient stability** (do they shift over time?)
3. **Input data distribution** (are new stores different?)
4. **Lambda appropriateness** (refit periodically)

**Recommendation:** Retrain monthly with new data

---

# Advanced: Ridge with Standardization

**Important:** Ridge is sensitive to scale!

```{r Part2_standardization_demo}
# glmnet standardizes by default
# To see unstandardized:
ridge_no_std <- glmnet(
  X_train, y_train, 
  alpha = 0,
  lambda = cv_ridge$lambda.1se,
  standardize = FALSE  # Don't standardize
)

# Compare coefficients
coef_comparison <- data.frame(
  Standardized = as.vector(coef(ridge_final))[-1],
  Unstandardized = as.vector(coef(ridge_no_std))[-1]
)

head(coef_comparison, 3)
```

**Always let glmnet standardize** (default behavior)!

---

# Ridge Formula Recap

**Mathematical formulation:**

$$\hat{\beta}^{\text{ridge}} = \underset{\beta}{\text{argmin}} \left\{ \sum_{i=1}^{n} (y_i - \beta_0 - \sum_{j=1}^{p} \beta_j x_{ij})^2 + \lambda \sum_{j=1}^{p} \beta_j^2 \right\}$$

**In plain English:**
- Find coefficients that minimize residuals (like OLS)
- But also keep coefficients small (penalty term)
- λ controls the trade-off between fit and simplicity

---

# Practical Tips for Ridge

**When fitting Ridge models:**

1. **Always use cross-validation** for λ selection
2. **Use lambda.1se** for better generalization (simpler model)
3. **Check coefficient paths** to understand behavior
4. **Standardize features** (glmnet does this automatically)
5. **Compare to OLS** on test set, not training set
6. **Test coefficient stability** across different samples

**Remember:** Goal is production performance, not training fit!

---

# What We've Learned

**Key Concepts:**
1. ✅ Ridge adds L2 penalty (sum of squared coefficients)
2. ✅ Lambda (λ) controls penalty strength
3. ✅ Cross-validation selects optimal λ
4. ✅ Ridge shrinks coefficients (but keeps all)
5. ✅ Coefficient paths show shrinkage process
6. ✅ Ridge improves test performance
7. ✅ Coefficients become more stable

**Business Value:**
- Better predictions on new data
- More reliable model in production
- Confidence in deployment

---

# Coming in Part 3: LASSO

**Ridge limitation:** Keeps ALL variables

**What if we want feature selection?**

**LASSO** (Least Absolute Shrinkage and Selection Operator):
- Sets some coefficients EXACTLY to zero
- Automatic feature selection
- Simpler, more interpretable models

**Preview:** From 10 features → 5-7 important features

---

# Summary: Part 2

**Ridge Regression:**
- Adds penalty for large coefficients
- Controlled by lambda (λ)
- Selected via cross-validation
- Improves generalization
- Stabilizes estimates

**Business Impact:**
- RetailCorp model now performs better on new stores
- Predictions are more reliable
- Ready for production testing

**Next:** Feature selection with LASSO!

---

class: inverse, center, middle

# 🎯 Classwork Time

## Building Ridge Models
### 25 minutes • 12 micro-exercises

You'll apply Ridge to housing price prediction

---

class: center, middle

# 10-Minute Break ☕

**When we return:** LASSO Regression (Part 3)

Feature selection and interpretability!

---

class: inverse, center, middle

# Part 3: LASSO Regression

## Feature Selection Through Regularization

---

# Where We Are Now

**Journey so far:**
- ✅ Part 1: Diagnosed overfitting and multicollinearity
- ✅ Part 2: Fixed overfitting with Ridge regression
- ❓ Ridge kept ALL 10 predictors (just shrunken)

**CFO's question:** "Which features actually matter for sales?"

**Ridge answer:** "All of them... sort of?"

**Today:** LASSO gives a clearer answer!

---

# The Ridge Limitation

**Problem with Ridge:**
- All coefficients are non-zero (just small)
- Can't say "this feature doesn't matter"
- Model still uses all 10 predictors

**Example Ridge coefficients:**
```
population_density:    0.023
median_income:         0.041
store_size_sqft:       0.0087
parking_spaces:        0.012
...all 10 features...  (small values)
```

**Business impact:** Hard to explain which features drive sales

---

# LASSO: The Key Difference

**LASSO** = Least Absolute Shrinkage and Selection Operator

**Ridge penalty (L2):**
$$\lambda \sum_{j=1}^{p} \beta_j^2$$

**LASSO penalty (L1):**
$$\lambda \sum_{j=1}^{p} |\beta_j|$$

**Critical difference:** LASSO can set coefficients **exactly to zero**!

---

# Why L1 Creates Sparsity

```{r Part3_lasso_geometry, echo=FALSE, fig.height=4.5}
# Conceptual visualization
theta <- seq(0, 2*pi, length = 100)

# L2 constraint (circle)
l2_x <- cos(theta)
l2_y <- sin(theta)

# L1 constraint (diamond)
l1_x <- c(1, 0, -1, 0, 1)
l1_y <- c(0, 1, 0, -1, 0)

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# Ridge (L2)
plot(l2_x, l2_y, type = "l", lwd = 2, col = "steelblue",
     xlim = c(-1.5, 1.5), ylim = c(-1.5, 1.5),
     xlab = "β₁", ylab = "β₂", main = "Ridge (L2): Circle")
abline(h = 0, v = 0, lty = 3, col = "gray")
points(0.7, 0.7, pch = 19, col = "red", cex = 1.5)
text(0.7, 0.7, "  Solution\n  (both non-zero)", pos = 4)

# LASSO (L1)
plot(l1_x, l1_y, type = "l", lwd = 2, col = "coral",
     xlim = c(-1.5, 1.5), ylim = c(-1.5, 1.5),
     xlab = "β₁", ylab = "β₂", main = "LASSO (L1): Diamond")
abline(h = 0, v = 0, lty = 3, col = "gray")
points(1, 0, pch = 19, col = "red", cex = 1.5)
text(1, 0, "  Solution\n  (β₂ = 0!)", pos = 4)

par(mfrow = c(1, 1))
```

**LASSO's diamond corners touch axes** → coefficients become exactly zero

---

# LASSO in One Equation

**LASSO optimization:**
$$\underset{\beta}{\text{minimize}} \left\{ \sum_{i=1}^{n} (y_i - \beta_0 - \sum_{j=1}^{p} \beta_j x_{ij})^2 + \lambda \sum_{j=1}^{p} |\beta_j| \right\}$$

**What it does:**
1. Minimize prediction error (first term)
2. Penalize absolute coefficient values (second term)
3. Result: Some coefficients → 0 (feature selection!)

**Lambda (λ) controls:** How many features to keep

---

# Fitting LASSO in R

```{r Part3_setup_data}
# Data already loaded from Part 2
# X_train, y_train, X_test, y_test already prepared

# Just verify they exist
cat("X_train dimensions:", dim(X_train), "\n")
cat("Ready to fit LASSO!\n")
```

**Note:** Using same data matrices as Ridge for fair comparison

---

# Our First LASSO Model

```{r Part3_first_lasso}
# Fit LASSO with lambda = 1000
lasso_basic <- glmnet(
  x = X_train,
  y = y_train,
  alpha = 1,      # alpha = 1 means LASSO (was 0 for Ridge)
  lambda = 1000
)

# Look at coefficients
coef(lasso_basic)
```

**Notice:** Some coefficients are EXACTLY zero!

---

# LASSO vs Ridge: Side by Side

```{r Part3_ridge_lasso_compare}
# Fit both with same lambda
ridge_compare <- glmnet(X_train, y_train, alpha = 0, lambda = 1000)
lasso_compare <- glmnet(X_train, y_train, alpha = 1, lambda = 1000)

# Compare coefficients
comparison <- tibble(
  Predictor = colnames(X_train),
  Ridge = as.vector(coef(ridge_compare))[-1],
  LASSO = as.vector(coef(lasso_compare))[-1],
  Ridge_NonZero = Ridge != 0,
  LASSO_NonZero = LASSO != 0
)

print(comparison)
```

---

# Coefficient Comparison Visualization

```{r Part3_coef_comparison, echo=FALSE, fig.height=4}
comparison %>%
  pivot_longer(c(Ridge, LASSO), names_to = "Model", values_to = "Coefficient") %>%
  mutate(Predictor = reorder(Predictor, abs(Coefficient))) %>%
  ggplot(aes(x = Predictor, y = Coefficient, fill = Model)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_hline(yintercept = 0, linetype = "solid") +
  coord_flip() +
  labs(title = "Ridge vs LASSO: Coefficient Comparison",
       subtitle = "LASSO sets some coefficients exactly to zero",
       x = NULL) +
  theme_minimal(base_size = 14)
```

**Key insight:** LASSO performs automatic feature selection!

---

# Cross-Validation for LASSO

```{r Part3_cv_lasso}
set.seed(42)
cv_lasso <- cv.glmnet(
  x = X_train,
  y = y_train,
  alpha = 1,        # LASSO
  nfolds = 10,
  type.measure = "mse"
)

# Optimal lambda values
cat("Lambda min:", cv_lasso$lambda.min, "\n")
cat("Lambda 1se:", cv_lasso$lambda.1se, "\n")
```

**Note:** LASSO typically needs smaller lambda than Ridge

---

# LASSO Cross-Validation Plot

```{r Part3_cv_plot, echo=FALSE, fig.height=4.5}
plot(cv_lasso, main = "LASSO Cross-Validation")
abline(v = log(cv_lasso$lambda.min), col = "red", lty = 2)
abline(v = log(cv_lasso$lambda.1se), col = "blue", lty = 2)

# Add text showing number of features at top
axis(3, at = log(cv_lasso$lambda), 
     labels = cv_lasso$nzero, 
     cex.axis = 0.7, col.axis = "darkgreen")
mtext("Number of Features", side = 3, line = 2.5, col = "darkgreen")

legend("topleft", 
       legend = c("Lambda min", "Lambda 1se"),
       col = c("red", "blue"), lty = 2)
```

**Top axis:** Shows how many features at each lambda!

---

# Understanding Feature Count

```{r Part3_feature_counts}
# How many features at different lambda values?
feature_counts <- data.frame(
  Lambda = cv_lasso$lambda,
  Features = cv_lasso$nzero,
  MSE = cv_lasso$cvm
) %>%
  filter(Lambda %in% c(cv_lasso$lambda.min, cv_lasso$lambda.1se) |
         Features %in% c(1, 5, 10))

print(feature_counts %>% head(5))
```

**Trade-off:** More features = lower MSE but more complex model

---

# Fitting Final LASSO Model

```{r Part3_final_lasso}
# Use lambda.1se (simpler model)
lasso_final <- glmnet(
  x = X_train,
  y = y_train,
  alpha = 1,
  lambda = cv_lasso$lambda.1se
)

# Extract coefficients
lasso_coefs <- coef(lasso_final)
print(lasso_coefs)

# Count non-zero coefficients
n_selected <- sum(lasso_coefs != 0) - 1  # -1 for intercept
cat("\nFeatures selected:", n_selected, "out of 10\n")
```

**LASSO automatically chose** ~5-7 important features!

---

# Which Features Survived?

```{r Part3_selected_features}
# Extract non-zero coefficients
selected_features <- data.frame(
  Feature = rownames(lasso_coefs)[-1],  # Remove intercept
  Coefficient = as.vector(lasso_coefs)[-1]
) %>%
  filter(Coefficient != 0) %>%
  arrange(desc(abs(Coefficient)))

print(selected_features)
```

**Business insight:** These are the key drivers of sales!

---

# Visualizing Selected Features

```{r Part3_selected_viz, echo=FALSE, fig.height=4}
selected_features %>%
  mutate(Feature = reorder(Feature, abs(Coefficient))) %>%
  ggplot(aes(x = Feature, y = Coefficient)) +
  geom_col(fill = "coral", width = 0.7) +
  geom_text(aes(label = round(Coefficient, 3)), 
            hjust = ifelse(selected_features$Coefficient > 0, -0.1, 1.1)) +
  coord_flip() +
  labs(title = "LASSO Selected Features",
       subtitle = paste("From 10 predictors to", n_selected, "key drivers"),
       x = NULL, y = "Coefficient Value") +
  theme_minimal(base_size = 14)
```

**CFO can now focus on these key features!**

---

# LASSO Performance: Training

```{r Part3_train_performance}
# Training predictions
lasso_train_pred <- predict(
  lasso_final, 
  newx = X_train,
  s = cv_lasso$lambda.1se
)

lasso_train_rmse <- sqrt(mean((y_train - lasso_train_pred)^2))

cat("OLS Training RMSE:   $", format(round(52143), big.mark = ","), "\n", sep = "")
cat("Ridge Training RMSE: $", format(round(53821), big.mark = ","), "\n", sep = "")
cat("LASSO Training RMSE: $", format(round(lasso_train_rmse), big.mark = ","), "\n", sep = "")
```

**Pattern:** More regularization → Higher training error (less overfitting)

---

# LASSO Performance: Testing

```{r Part3_test_performance}
# Test predictions
lasso_test_pred <- predict(
  lasso_final,
  newx = X_test,
  s = cv_lasso$lambda.1se
)

lasso_test_rmse <- sqrt(mean((y_test - lasso_test_pred)^2))

cat("OLS Test RMSE:   $", format(round(61247), big.mark = ","), "\n", sep = "")
cat("Ridge Test RMSE: $", format(round(58634), big.mark = ","), "\n", sep = "")
cat("LASSO Test RMSE: $", format(round(lasso_test_rmse), big.mark = ","), "\n", sep = "")
```

**Key result:** LASSO performs similarly to Ridge BUT simpler!

---

# Three-Way Performance Comparison

```{r Part3_three_way_comparison, echo=FALSE, fig.height=4.5}
performance_data <- tibble(
  Model = rep(c("OLS", "Ridge", "LASSO"), each = 2),
  Dataset = rep(c("Training", "Test"), 3),
  RMSE = c(52143, 61247,  # OLS
           53821, 58634,  # Ridge
           lasso_train_rmse, lasso_test_rmse)  # LASSO
)

ggplot(performance_data, aes(x = Dataset, y = RMSE, fill = Model)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(aes(label = paste("$", format(round(RMSE/1000), big.mark = ","), "K", sep = "")),
            position = position_dodge(width = 0.7), vjust = -0.5, size = 3) +
  labs(title = "OLS vs Ridge vs LASSO: Performance Comparison",
       subtitle = "LASSO: Similar performance to Ridge, but simpler (fewer features)",
       y = "RMSE") +
  scale_y_continuous(labels = scales::dollar) +
  theme_minimal(base_size = 14)
```

---

# The LASSO Coefficient Path

```{r Part3_coef_path}
# Fit LASSO across many lambda values
lasso_path <- glmnet(
  x = X_train,
  y = y_train,
  alpha = 1,
  lambda = 10^seq(5, -2, length = 100)
)

# Plot the path
plot(lasso_path, xvar = "lambda", label = TRUE,
     main = "LASSO Coefficient Paths")
abline(v = log(cv_lasso$lambda.1se), lty = 2, col = "red")
```

**Each line is a predictor** - watch them go to zero!

---

# Coefficient Path Interpretation

```{r Part3_path_detail, echo=FALSE, fig.height=4.5}
plot(lasso_path, xvar = "lambda", label = TRUE,
     main = "LASSO: Features Drop Out as Lambda Increases")
abline(v = log(cv_lasso$lambda.1se), lty = 2, col = "red", lwd = 2)
abline(v = log(cv_lasso$lambda.min), lty = 2, col = "blue", lwd = 2)

legend("topright", 
       legend = c("Lambda 1se", "Lambda min", "Feature path"),
       col = c("red", "blue", "black"), lty = c(2, 2, 1), lwd = c(2, 2, 1))
```

**Pattern:**
- Left: All features active
- Middle: Some drop to zero (optimal region)
- Right: Most/all features zero

---

# Lambda and Feature Selection

```{r Part3_lambda_features}
# Test several lambda values
test_lambdas <- c(10, 100, 1000, 10000)

lambda_results <- map_dfr(test_lambdas, function(lam) {
  model <- glmnet(X_train, y_train, alpha = 1, lambda = lam)
  n_features <- sum(coef(model) != 0) - 1
  
  test_pred <- predict(model, newx = X_test)
  test_rmse <- sqrt(mean((y_test - test_pred)^2))
  
  tibble(Lambda = lam, Features = n_features, Test_RMSE = test_rmse)
})

print(lambda_results)
```

---

# Lambda-Features-Performance Curve

```{r Part3_lambda_curve, echo=FALSE, fig.height=4}
lambda_results %>%
  ggplot(aes(x = Features, y = Test_RMSE)) +
  geom_line(size = 1.5, color = "coral") +
  geom_point(size = 3, color = "coral") +
  geom_text(aes(label = paste("λ =", Lambda)), 
            vjust = -0.5, size = 3) +
  labs(title = "LASSO: Features vs Performance",
       subtitle = "Sweet spot around 5-7 features",
       x = "Number of Features Selected",
       y = "Test RMSE") +
  scale_y_continuous(labels = scales::dollar) +
  theme_minimal(base_size = 14)
```

**Optimal:** Balance between simplicity and accuracy

---

# Interpreting Feature Importance

```{r Part3_feature_importance}
# Order features by absolute coefficient value
importance <- data.frame(
  Feature = rownames(lasso_coefs)[-1],
  Coefficient = as.vector(lasso_coefs)[-1],
  Importance = abs(as.vector(lasso_coefs)[-1])
) %>%
  arrange(desc(Importance)) %>%
  mutate(
    Selected = Coefficient != 0,
    Rank = row_number()
  )

print(importance)
```

---

# Feature Importance Visualization

```{r Part3_importance_viz, echo=FALSE, fig.height=4.5}
importance %>%
  mutate(Feature = reorder(Feature, Importance)) %>%
  ggplot(aes(x = Feature, y = Importance, fill = Selected)) +
  geom_col(width = 0.7) +
  coord_flip() +
  labs(title = "Feature Importance from LASSO",
       subtitle = "Dark bars = Selected features, Light bars = Dropped",
       x = NULL, y = "Absolute Coefficient Value") +
  scale_fill_manual(values = c("FALSE" = "gray80", "TRUE" = "coral"),
                    labels = c("Dropped", "Selected")) +
  theme_minimal(base_size = 14)
```

**Clear story for stakeholders!**

---

# Business Interpretation

**What we tell the CFO:**

"Our LASSO model identified **[5-7] key drivers** of store sales:

**Top 3 Most Important:**
1. [Feature 1] - Each unit increase adds $X to sales
2. [Feature 2] - Critical for store performance
3. [Feature 3] - Strong positive impact

**Dropped features:** [List 3-5 features with zero coefficients]

**Business value:**
- ✅ Simpler model (easier to explain and monitor)
- ✅ Focus resources on key features
- ✅ Similar accuracy to complex model
- ✅ Clear priorities for new store planning"

---

# LASSO for Different Business Questions

**Q1: "Which features matter most?"**
→ Use LASSO, look at non-zero coefficients

**Q2: "Can we reduce data collection costs?"**
→ Use LASSO, only measure selected features

**Q3: "What should we focus on for new stores?"**
→ Use LASSO features as checklist

**Q4: "Is this feature worth the measurement cost?"**
→ If LASSO drops it, probably not!

---

# Ridge vs LASSO: When to Use Each?

**Use Ridge when:**
- ✅ All features might be relevant
- ✅ Features are highly correlated
- ✅ Prediction accuracy is priority
- ✅ Stakeholders want all information

**Use LASSO when:**
- ✅ Need feature selection
- ✅ Want interpretable model
- ✅ Many features, limited budget
- ✅ Stakeholders want "top drivers"

**Both are good for:** Multicollinearity and overfitting

---

# Stability of Feature Selection

```{r Part3_stability_test}
# Fit LASSO on 3 different samples
stability_results <- map_dfr(1:3, function(i) {
  idx <- sample(1:nrow(X_train), size = 0.9 * nrow(X_train))
  
  cv_temp <- cv.glmnet(X_train[idx, ], y_train[idx], 
                       alpha = 1, nfolds = 5)
  model <- glmnet(X_train[idx, ], y_train[idx], 
                  alpha = 1, lambda = cv_temp$lambda.1se)
  
  tibble(
    Sample = i,
    Feature = colnames(X_train),
    Selected = as.vector(coef(model))[-1] != 0
  )
})

# Which features consistently selected?
stability_summary <- stability_results %>%
  group_by(Feature) %>%
  summarise(Times_Selected = sum(Selected))

print(stability_summary)
```

---

# Feature Selection Stability

```{r Part3_stability_viz, echo=FALSE, fig.height=4}
stability_summary %>%
  mutate(Feature = reorder(Feature, Times_Selected)) %>%
  ggplot(aes(x = Feature, y = Times_Selected)) +
  geom_col(aes(fill = Times_Selected == 3), width = 0.7) +
  geom_hline(yintercept = 2, linetype = "dashed", color = "red") +
  coord_flip() +
  scale_y_continuous(breaks = 0:3) +
  labs(title = "LASSO Feature Selection Stability",
       subtitle = "Selected in how many of 3 samples?",
       x = NULL, y = "Times Selected (out of 3)") +
  scale_fill_manual(values = c("coral", "darkgreen"),
                    labels = c("Unstable", "Stable")) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

**Stable features:** Selected every time (most reliable!)

---

# Practical Considerations

**LASSO in production:**

1. **Feature engineering:** Create good candidate features first
2. **Standardization:** Always standardize (glmnet does this)
3. **Lambda selection:** Use lambda.1se for stability
4. **Feature stability:** Verify on multiple samples
5. **Business logic:** Don't drop features that make business sense
6. **Monitoring:** Track which features stay selected over time

---

# Saving LASSO Models

```{r Part3_save_model, eval=FALSE}
# Save for production
lasso_artifacts <- list(
  model = lasso_final,
  lambda = cv_lasso$lambda.1se,
  selected_features = selected_features$Feature,
  performance = list(
    train_rmse = lasso_train_rmse,
    test_rmse = lasso_test_rmse
  )
)

saveRDS(lasso_artifacts, "models/lasso_sales_model.rds")

# Later: Load and predict
artifacts <- readRDS("models/lasso_sales_model.rds")
predictions <- predict(artifacts$model, 
                      newx = new_data_matrix, 
                      s = artifacts$lambda)
```

**Pro tip:** Save selected features list for documentation!

---

# Advanced: Adaptive LASSO

```{r Part3_adaptive_lasso, eval=FALSE}
# Stage 1: Get initial estimates (Ridge or OLS)
ridge_init <- cv.glmnet(X_train, y_train, alpha = 0)
init_coef <- coef(ridge_init, s = "lambda.1se")[-1]

# Stage 2: Weight LASSO penalty by initial estimates
penalty_weights <- 1 / abs(init_coef)

# Fit adaptive LASSO
adaptive_lasso <- cv.glmnet(
  X_train, y_train,
  alpha = 1,
  penalty.factor = penalty_weights
)
```

**Benefit:** Less bias, better feature selection

---

# Comparing All Three Methods

```{r Part3_final_comparison}
# Summary table
comparison_table <- tibble(
  Method = c("OLS", "Ridge", "LASSO"),
  Features = c(10, 10, n_selected),
  Train_RMSE = c(52143, 53821, lasso_train_rmse),
  Test_RMSE = c(61247, 58634, lasso_test_rmse),
  Interpretability = c("Medium", "Low", "High")
)

print(comparison_table)
```

**Key insights:**
- OLS: Overfits (worst test error)
- Ridge: Good performance, all features
- LASSO: Similar to Ridge, fewer features (winner for interpretability!)

---

# What We've Learned

**Key Concepts:**
1. ✅ LASSO adds L1 penalty (sum of absolute coefficients)
2. ✅ L1 penalty creates exact zeros (feature selection)
3. ✅ Cross-validation selects optimal lambda
4. ✅ Coefficient paths show feature dropout
5. ✅ Fewer features = simpler, more interpretable
6. ✅ Similar performance to Ridge

**Business Value:**
- Clear answer to "what matters?"
- Simpler models for stakeholders
- Focus resources on key drivers
- Data collection cost savings

---

# Ridge vs LASSO Summary

| Aspect | Ridge (L2) | LASSO (L1) |
|--------|-----------|------------|
| **Penalty** | Sum of squared coefficients | Sum of absolute coefficients |
| **Feature selection** | No (all kept) | Yes (sets some to zero) |
| **Coefficients** | All shrunk | Some zero, some shrunk |
| **Use when** | All features relevant | Want interpretability |
| **Interpretability** | Lower | Higher |
| **Performance** | Slightly better | Similar |
| **Alpha in glmnet** | 0 | 1 |

---

# Coming in Part 4: Elastic Net

**Question:** "Can we get the best of both worlds?"

**Answer:** Elastic Net!

**Elastic Net penalty:**
$$\lambda \left[ \alpha |\beta_j| + (1-\alpha) \beta_j^2 \right]$$

- α = 0: Pure Ridge
- α = 1: Pure LASSO  
- 0 < α < 1: Combination!

**Benefits:**
- Feature selection (like LASSO)
- Handles correlated features well (like Ridge)
- Often better than either alone

---

# Summary: Part 3

**LASSO Regression:**
- L1 penalty for feature selection
- Sets coefficients exactly to zero
- Simpler, more interpretable models
- Similar performance to Ridge

**RetailCorp Application:**
- Identified 5-7 key sales drivers
- Dropped 3-5 less important features
- Similar accuracy, easier to explain
- Clear priorities for operations team

**Next:** Combine Ridge + LASSO = Elastic Net!

---

class: inverse, center, middle

# 🎯 Classwork Time

## LASSO Feature Selection
### 25 minutes • 12 micro-exercises

Apply LASSO to housing prices and see which features matter!

---

class: center, middle

# 10-Minute Break ☕

**When we return:** Elastic Net (Part 4)

The best of both Ridge and LASSO!

---
---

---

# Part 4: Elastic Net
## Slides 91-120

---

class: inverse, center, middle

# Elastic Net

## The Best of Both Worlds

---

# Where We Are Now

**Journey so far:**
- ✅ Part 1: Diagnosed overfitting and multicollinearity
- ✅ Part 2: Ridge shrinks all coefficients (keeps all features)
- ✅ Part 3: LASSO selects features (sets some to zero)
- ❓ Can we combine Ridge + LASSO strengths?

**The Problem:**
- Ridge: Great with correlated features, but keeps everything
- LASSO: Great feature selection, but struggles with correlation

**Today:** Elastic Net gives us both!

---

# The Motivation for Elastic Net

**Real-world scenario at RetailCorp:**

We have correlated features:
- `square_feet` and `lot_size` are highly correlated
- `bedrooms` and `bathrooms` correlate
- `income` and `property_value` correlate

**Ridge:** Keeps all features, hard to interpret (10 features)

**LASSO:** Randomly picks one from each pair, unstable results

**Elastic Net:** Keeps both from correlated pairs, still does selection!

---

# The Elastic Net Formula

**Elastic Net combines both penalties:**

$$\underset{\beta}{\text{minimize}} \left\{ \sum_{i=1}^{n} (y_i - \hat{y}_i)^2 + \lambda \left[ \frac{1-\alpha}{2} \sum_{j=1}^{p} \beta_j^2 + \alpha \sum_{j=1}^{p} |\beta_j| \right] \right\}$$

**Two hyperparameters:**
- **λ (lambda):** Overall penalty strength (like before)
- **α (alpha):** Mix between Ridge and LASSO
  - α = 0: Pure Ridge (L2 only)
  - α = 1: Pure LASSO (L1 only)
  - 0 < α < 1: Elastic Net mixture (both L1 + L2)

---

# Understanding Alpha (α)

```{r Part4_alpha_visual, echo=FALSE, fig.height=4.5}
# Visualize what alpha controls
alpha_seq <- seq(0, 1, by = 0.1)

alpha_data <- tibble(
  Alpha = alpha_seq,
  L2_Weight = 1 - alpha_seq,
  L1_Weight = alpha_seq
)

alpha_data %>%
  pivot_longer(c(L2_Weight, L1_Weight), names_to = "Penalty", values_to = "Weight") %>%
  ggplot(aes(x = Alpha, y = Weight, color = Penalty)) +
  geom_line(size = 2) +
  geom_point(size = 3) +
  annotate("text", x = 0, y = 0.5, label = "Ridge\n(L2 only)", 
           size = 5, hjust = -0.2) +
  annotate("text", x = 1, y = 0.5, label = "LASSO\n(L1 only)", 
           size = 5, hjust = 1.2) +
  annotate("rect", xmin = 0.3, xmax = 0.7, ymin = 0, ymax = 1,
           alpha = 0.1, fill = "green") +
  annotate("text", x = 0.5, y = 0.9, label = "Elastic Net\nSweet Spot", 
           color = "darkgreen", size = 5) +
  labs(title = "Alpha Controls the Ridge-LASSO Mix",
       subtitle = "Alpha = 0: Pure Ridge | Alpha = 1: Pure LASSO | 0 < Alpha < 1: Combination",
       y = "Penalty Weight") +
  scale_color_manual(values = c("steelblue", "coral"),
                     labels = c("L1 (LASSO)", "L2 (Ridge)")) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")
```

**Sweet spot:** Usually α between 0.3 and 0.7

---

# Why Elastic Net Helps

**Problem: LASSO with correlated features**

Imagine `square_feet` and `lot_size` are highly correlated (r = 0.85):

- **LASSO:** Randomly picks `square_feet` OR `lot_size`
  - Unstable across different samples
  - Changes which feature is selected
  
- **Elastic Net:** Keeps BOTH `square_feet` AND `lot_size`
  - More stable
  - Recognizes both are useful
  - Groups correlated features together

**This is called the "grouping effect"**

---

# Fitting Elastic Net in R

```{r Part4_enet_basic}
# Data already prepared from Part 3
# X_train, y_train, X_test, y_test are ready

# Fit Elastic Net with alpha = 0.5 (50-50 mix)
set.seed(123)
cv_enet <- cv.glmnet(
  x = X_train,
  y = y_train,
  alpha = 0.5,      # Elastic Net (halfway between Ridge and LASSO)
  nfolds = 10,
  type.measure = "mse"
)

# Optimal lambda values
cat("Lambda min:", cv_enet$lambda.min, "\n")
cat("Lambda 1se:", cv_enet$lambda.1se, "\n")
```

**Simple:** Just change `alpha` from 0 (Ridge) or 1 (LASSO) to 0.5!

---

# Cross-Validation Plot

```{r Part4_cv_plot, echo=FALSE, fig.height=4.5}
plot(cv_enet, main = "Elastic Net Cross-Validation (α = 0.5)")
abline(v = log(cv_enet$lambda.min), col = "red", lty = 2, lwd = 2)
abline(v = log(cv_enet$lambda.1se), col = "blue", lty = 2, lwd = 2)

# Add feature count on top
axis(3, at = log(cv_enet$lambda), 
     labels = cv_enet$nzero, 
     cex.axis = 0.7, col.axis = "darkgreen")
mtext("Number of Features", side = 3, line = 2.5, col = "darkgreen")

legend("topleft", 
       legend = c("Lambda min", "Lambda 1se"),
       col = c("red", "blue"), lty = 2, lwd = 2)
```

**Top axis:** Shows how many features selected at each lambda

---

# Fitting Final Elastic Net Model

```{r Part4_final_model}
# Fit final model using lambda.1se
enet_final <- glmnet(
  x = X_train,
  y = y_train,
  alpha = 0.5,
  lambda = cv_enet$lambda.1se
)

# Extract coefficients
enet_coefs <- coef(enet_final)
print(enet_coefs)

# Count non-zero coefficients
n_selected_enet <- sum(enet_coefs != 0) - 1  # -1 for intercept
cat("\nElastic Net selected:", n_selected_enet, "features out of 10\n")
```

---

# Which Features Did Elastic Net Select?

```{r Part4_selected_features}
# Extract selected features
enet_selected <- data.frame(
  Feature = rownames(enet_coefs)[-1],
  Coefficient = as.vector(enet_coefs)[-1]
) %>%
  filter(Coefficient != 0) %>%
  arrange(desc(abs(Coefficient)))

print(enet_selected)
```

**Compare:** Similar to LASSO but potentially more stable

---

# Comparing All Three Methods

```{r Part4_three_way_compare}
# Get coefficients from all three regularization methods
ridge_coefs_all <- as.vector(coef(ridge_final))[-1]
lasso_coefs_all <- as.vector(coef(lasso_final))[-1]
enet_coefs_all <- as.vector(coef(enet_final))[-1]

# Create comparison
all_three <- data.frame(
  Feature = colnames(X_train),
  Ridge = ridge_coefs_all,
  LASSO = lasso_coefs_all,
  Elastic_Net = enet_coefs_all
) %>%
  mutate(
    Ridge_NonZero = Ridge != 0,
    LASSO_NonZero = LASSO != 0,
    Enet_NonZero = Elastic_Net != 0
  )

# Show selection summary
cat("Ridge selected:", sum(all_three$Ridge_NonZero), "\n")
cat("LASSO selected:", sum(all_three$LASSO_NonZero), "\n")
cat("Elastic Net selected:", sum(all_three$Enet_NonZero), "\n")
```

---

# Three-Way Visualization

```{r Part4_three_viz, echo=FALSE, fig.height=4.5}
all_three %>%
  select(Feature, Ridge, LASSO, Elastic_Net) %>%
  pivot_longer(-Feature, names_to = "Method", values_to = "Coefficient") %>%
  mutate(Feature = reorder(Feature, abs(Coefficient))) %>%
  ggplot(aes(x = Feature, y = Coefficient, fill = Method)) +
  geom_col(position = "dodge", width = 0.8) +
  geom_hline(yintercept = 0) +
  coord_flip() +
  labs(title = "Coefficient Comparison: Ridge vs LASSO vs Elastic Net",
       subtitle = "Elastic Net balances between Ridge (keeps all) and LASSO (drops some)",
       x = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")
```

---

# Elastic Net Performance

```{r Part4_performance}
# Training predictions
enet_train_pred <- predict(enet_final, newx = X_train, s = cv_enet$lambda.1se)
enet_train_rmse <- sqrt(mean((y_train - enet_train_pred)^2))

# Test predictions
enet_test_pred <- predict(enet_final, newx = X_test, s = cv_enet$lambda.1se)
enet_test_rmse <- sqrt(mean((y_test - enet_test_pred)^2))

# Display all results
cat("OLS Test RMSE:         $", format(round(61247), big.mark = ","), "\n", sep = "")
cat("Ridge Test RMSE:       $", format(round(ridge_test_rmse), big.mark = ","), "\n", sep = "")
cat("LASSO Test RMSE:       $", format(round(lasso_test_rmse), big.mark = ","), "\n", sep = "")
cat("Elastic Net Test RMSE: $", format(round(enet_test_rmse), big.mark = ","), "\n", sep = "")
```

**Typically:** Elastic Net performs similar to or slightly better than best of Ridge/LASSO

---

# Complete Performance Comparison

```{r Part4_perf_viz, echo=FALSE, fig.height=4.5}
perf_data <- tibble(
  Model = rep(c("OLS", "Ridge", "LASSO", "Elastic Net"), each = 2),
  Dataset = rep(c("Training", "Test"), 4),
  RMSE = c(52143, 61247,                      # OLS
           ridge_train_rmse, ridge_test_rmse,  # Ridge
           lasso_train_rmse, lasso_test_rmse,  # LASSO
           enet_train_rmse, enet_test_rmse)    # Elastic Net
)

ggplot(perf_data, aes(x = Dataset, y = RMSE, fill = Model)) +
  geom_col(position = "dodge", width = 0.8) +
  geom_text(aes(label = paste0("$", format(round(RMSE/1000), big.mark = ","), "K")),
            position = position_dodge(width = 0.8), vjust = -0.5, size = 3) +
  labs(title = "Complete Model Comparison",
       subtitle = "All regularization methods reduce overfitting compared to OLS",
       y = "RMSE") +
  scale_y_continuous(labels = scales::dollar) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")
```

---

# Tuning Alpha: Finding Best Mix

**So far we used α = 0.5, but is that optimal?**

Let's test multiple alpha values:

```{r Part4_tune_alpha}
# Test different alpha values
alpha_values <- seq(0, 1, by = 0.2)

alpha_results <- map_dfr(alpha_values, function(a) {
  set.seed(123)
  cv_temp <- cv.glmnet(X_train, y_train, alpha = a, nfolds = 10)
  
  # Fit with lambda.1se
  model <- glmnet(X_train, y_train, alpha = a, lambda = cv_temp$lambda.1se)
  test_pred <- predict(model, newx = X_test, s = cv_temp$lambda.1se)
  test_rmse <- sqrt(mean((y_test - test_pred)^2))
  
  # Count features
  n_features <- sum(coef(model) != 0) - 1
  
  tibble(Alpha = a, Test_RMSE = test_rmse, Features = n_features)
})

print(alpha_results)
```

---

# Alpha Tuning Visualization

```{r Part4_alpha_tune_viz, echo=FALSE, fig.height=4.5}
ggplot(alpha_results, aes(x = Alpha)) +
  geom_line(aes(y = Test_RMSE), color = "steelblue", size = 1.5) +
  geom_point(aes(y = Test_RMSE), color = "steelblue", size = 3) +
  geom_text(aes(y = Test_RMSE, label = Features), 
            vjust = -1, color = "coral", size = 3) +
  annotate("text", x = 0.5, y = max(alpha_results$Test_RMSE), 
           label = "Numbers = Features selected", 
           color = "coral", hjust = 0) +
  labs(title = "Effect of Alpha on Test Performance",
       subtitle = "Lower RMSE = better performance | Numbers show features selected",
       x = "Alpha (0 = Ridge, 1 = LASSO)",
       y = "Test RMSE") +
  scale_y_continuous(labels = scales::dollar) +
  theme_minimal(base_size = 14)
```

---

# Finding Optimal Alpha

```{r Part4_best_alpha}
# Which alpha gives best test performance?
best_alpha_row <- alpha_results %>%
  arrange(Test_RMSE) %>%
  slice(1)

cat("Best alpha:", best_alpha_row$Alpha, "\n")
cat("Test RMSE: $", format(round(best_alpha_row$Test_RMSE), big.mark = ","), "\n", sep = "")
cat("Features selected:", best_alpha_row$Features, "\n")
```

**Insight:** Optimal alpha usually between 0.2 and 0.8 for real data

---

# When to Use Elastic Net

**Choose Elastic Net when:**

✅ **Features are correlated** (common in real data)  
✅ **Want feature selection** (interpretability)  
✅ **Need stability** (consistent results)  
✅ **Unsure between Ridge and LASSO** (get both benefits)

**Example scenarios:**
- Genomics: Gene pathways are correlated
- Finance: Economic indicators correlate
- Marketing: Channel metrics correlate
- Real estate: Building features correlate (our case!)

---

# Ridge vs LASSO vs Elastic Net Summary

| Aspect | Ridge (α=0) | LASSO (α=1) | Elastic Net (0<α<1) |
|--------|-------------|-------------|---------------------|
| **Penalty** | L2 only | L1 only | L1 + L2 mix |
| **Feature selection** | No (keeps all) | Yes (sets to zero) | Yes (sets to zero) |
| **Correlated features** | Keeps all | Picks one randomly | Keeps group |
| **Stability** | High | Medium | High |
| **Interpretability** | Low | High | High |
| **Best for** | All relevant | Many irrelevant | Correlated groups |

---

# Final Business Comparison

```{r Part4_business_table}
# Create business comparison
biz_comparison <- tibble(
  Method = c("OLS", "Ridge", "LASSO", "Elastic Net"),
  Features = c(10, 10, n_selected, n_selected_enet),
  Test_RMSE = round(c(61247, ridge_test_rmse, lasso_test_rmse, enet_test_rmse)),
  Interpretability = c("Medium", "Low", "High", "High"),
  Stability = c("Low", "High", "Medium", "High"),
  Recommendation = c("❌ Overfits", "✅ Good", "✅ Good", "✅ Best")
)

kable(biz_comparison, 
      caption = "Business Decision Matrix",
      format.args = list(big.mark = ","))
```

---

# Practical Recommendation for RetailCorp

**Final Recommendation: Use Elastic Net**

**Rationale:**
1. ✅ Test RMSE of $X,XXX (similar to Ridge/LASSO)
2. ✅ Selects X-X key features (interpretable for CFO)
3. ✅ Handles correlations in store characteristics
4. ✅ More stable than LASSO alone
5. ✅ Production-ready confidence

**Key sales drivers identified:**
- Square footage, bedrooms, bathrooms
- School quality, neighborhood safety
- Location accessibility

**Implementation:** Easy with glmnet, production-ready

---

# Saving Elastic Net Model

```{r Part4_save, eval=FALSE}
# Save complete model artifacts
enet_artifacts <- list(
  model = enet_final,
  alpha = 0.5,
  lambda = cv_enet$lambda.1se,
  selected_features = enet_selected$Feature,
  coefficients = enet_coefs,
  performance = list(
    train_rmse = enet_train_rmse,
    test_rmse = enet_test_rmse
  ),
  tuning_results = alpha_results
)

saveRDS(enet_artifacts, "models/elastic_net_final.rds")

# Load later
loaded <- readRDS("models/elastic_net_final.rds")
new_pred <- predict(loaded$model, newx = new_data, s = loaded$lambda)
```

---

# What We've Learned

**Key Concepts:**
1. ✅ Elastic Net combines Ridge (L2) + LASSO (L1)
2. ✅ Alpha (α) controls the L1/L2 mix
3. ✅ Lambda (λ) controls overall penalty strength
4. ✅ Handles correlated features better than LASSO
5. ✅ Still performs feature selection (unlike Ridge)
6. ✅ More stable than LASSO, more selective than Ridge

**Business Value:**
- Best-of-both-worlds approach
- Works with realistic correlated data
- Interpretable AND powerful
- Production-ready stability

---

# Complete Regularization Summary

**The Complete Journey:**

**Part 1:** Diagnosed the problem
- Overfitting and multicollinearity

**Part 2:** Ridge Regression (α = 0)
- Shrinks all coefficients, keeps all features

**Part 3:** LASSO Regression (α = 1)
- Sets some to zero, automatic feature selection

**Part 4:** Elastic Net (0 < α < 1)
- Combines both penalties, balanced approach

---

# Practical Tips for Using Elastic Net

**1. Start with α = 0.5**
- Good default (50-50 mix)
- Works well in practice

**2. Tune if needed**
- Test α = 0, 0.2, 0.4, 0.6, 0.8, 1.0
- Pick best via cross-validation

**3. Always use cross-validation**
- For lambda selection
- Use lambda.1se for stability

**4. Check feature stability**
- Refit on bootstrap samples
- Verify consistent selection

**5. Document everything**
- Save alpha and lambda values
- Record selected features
- Track performance metrics

---

# Industry Applications

**Where Elastic Net shines:**

**Healthcare:**
- Patient features correlate (age, BMI, conditions)
- Need interpretable risk models
- **Solution:** Elastic Net stable + selective

**Finance:**
- Economic indicators correlate
- Need robust trading signals
- **Solution:** Elastic Net groups correlated factors

**Marketing:**
- Channel metrics correlate (impressions, clicks, reach)
- Need budget allocation
- **Solution:** Elastic Net identifies key channels

**Real Estate (Our case!):**
- Building features correlate
- Need pricing insights
- **Solution:** Elastic Net handles correlation + selection

---

# Limitations to Remember

**What regularization CAN'T fix:**

❌ **Poor quality data**
- Garbage in, garbage out

❌ **Wrong features**
- Can't create signal from noise

❌ **Non-linear relationships**
- Need feature engineering or other models

❌ **Causal inference**
- Correlation ≠ causation

❌ **Extrapolation**
- Still unreliable outside training range

**Regularization helps with:** Overfitting and multicollinearity ONLY

---

# Summary: Part 4

**Elastic Net:**
- Combines L1 (LASSO) + L2 (Ridge) penalties
- Controlled by two hyperparameters: alpha (α) and lambda (λ)
- Handles correlated features while doing feature selection
- More stable than LASSO alone
- More interpretable than Ridge alone
- Usually optimal alpha between 0.3 and 0.7

**RetailCorp Recommendation:**
- Use Elastic Net with α = 0.5 (or tune α)
- Identifies 8-9 key sales drivers
- Test RMSE competitive with Ridge/LASSO
- Better stability for production deployment

---

class: inverse, center, middle

# 🎯 Classwork Time

## Elastic Net Practice
### 20 minutes • 8 micro-exercises

Implement Elastic Net and tune alpha on housing data!

---

class: center, middle

# 10-Minute Break ☕

**When we return:** Advanced Topics (Part 5)

caret integration and production workflows!

---
---

# Part 5: Advanced Topics & Practical Implementation
## Slides 121-140

---

class: inverse, center, middle

# Advanced Topics

## caret Integration & Decision Framework

---

# Where We Are Now

**Complete journey:**
- ✅ Part 1: Diagnosed overfitting and multicollinearity
- ✅ Part 2: Ridge regression (L2 penalty)
- ✅ Part 3: LASSO regression (L1 penalty, feature selection)
- ✅ Part 4: Elastic Net (combined L1 + L2)
- 🎯 **Today:** Professional workflows with caret

**Why caret?**
- Unified interface for all models
- Automated hyperparameter tuning
- Standardized cross-validation
- Production-ready pipelines

---

# The caret Package

**caret** = Classification And Regression Training

```{r Part5_caret_intro}
library(caret)

# caret provides 238+ models with unified interface
# All use same syntax: train(x, y, method = "...")

# Models available:
# - "lm" (OLS)
# - "glmnet" (Ridge/LASSO/Elastic Net)
# - "rf" (Random Forest)
# - "xgbTree" (XGBoost)
# - and 234+ more!
```

**Key benefit:** Learn one syntax, use any model!

---

# Basic caret Workflow

```{r Part5_caret_basic, eval=FALSE}
# 1. Define training control
train_control <- trainControl(
  method = "cv",          # Cross-validation
  number = 10,            # 10 folds
  verboseIter = FALSE     # Suppress progress
)

# 2. Train model
model <- train(
  x = X_train,
  y = y_train,
  method = "glmnet",      # Model type
  trControl = train_control
)

# 3. Predict
predictions <- predict(model, newdata = X_test)
```

**Three steps:** Control → Train → Predict

---

# Setting Up Training Control

```{r Part5_train_control}
# Define cross-validation strategy
train_control <- trainControl(
  method = "cv",                    # Cross-validation
  number = 10,                      # 10-fold CV
  search = "grid",                  # Grid search for tuning
  savePredictions = "final",        # Save predictions
  classProbs = FALSE,               # Regression (not classification)
  summaryFunction = defaultSummary  # RMSE, R-squared, MAE
)

print(train_control)
```

**trainControl** defines HOW to train and validate

---

# Defining the Tuning Grid

```{r Part5_tuning_grid}
# Create grid for alpha and lambda
tune_grid <- expand.grid(
  alpha = seq(0, 1, by = 0.1),              # 0 to 1 in steps of 0.1
  lambda = 10^seq(5, -2, length.out = 50)   # 50 lambda values
)

# See first few rows
head(tune_grid)

# Total combinations
cat("Total combinations to test:", nrow(tune_grid), "\n")
```

**Grid search:** Tests all combinations of alpha × lambda

---

# Training with caret

```{r Part5_train_caret}
# Train Elastic Net with caret
set.seed(123)

enet_caret <- train(
  x = X_train,
  y = y_train,
  method = "glmnet",
  trControl = train_control,
  tuneGrid = tune_grid,
  metric = "RMSE"  # Optimize for lowest RMSE
)

# This may take 1-2 minutes...
print(enet_caret)
```

**caret automatically:**
- Tests all alpha/lambda combinations
- Performs 10-fold CV for each
- Selects best parameters

---

# Best Parameters Found

```{r Part5_best_params}
# Extract best hyperparameters
best_params <- enet_caret$bestTune
cat("Best alpha:", best_params$alpha, "\n")
cat("Best lambda:", best_params$lambda, "\n")

# Get best model performance
best_results <- enet_caret$results %>%
  filter(alpha == best_params$alpha, lambda == best_params$lambda)

cat("\nBest CV RMSE:", round(best_results$RMSE), "\n")
cat("Best R-squared:", round(best_results$Rsquared, 3), "\n")
```

**Optimal values:** Found through exhaustive search

---

# Visualizing the Tuning Process

```{r Part5_plot_tuning, fig.height=4.5}
# Plot RMSE across parameter space
plot(enet_caret, 
     plotType = "level",  # Heatmap
     main = "RMSE Across Alpha-Lambda Grid")

# Can also use plotType = "scatter"
```

**Heatmap shows:** Which alpha/lambda combinations work best

---

# Alternative: Line Plot

```{r Part5_plot_lines, fig.height=4.5}
# Line plot for each alpha
plot(enet_caret, 
     plotType = "scatter",
     main = "RMSE vs Lambda for Different Alpha Values")
```

**Each line:** Different alpha value (0 = Ridge, 1 = LASSO)

---

# Variable Importance

```{r Part5_var_importance}
# Extract variable importance
importance <- varImp(enet_caret, scale = TRUE)
print(importance)

# Plot importance
plot(importance, 
     top = 10,
     main = "Top 10 Most Important Features")
```

**Importance:** Based on absolute coefficient values

---

# Making Predictions

```{r Part5_predictions_caret}
# Predict on test set
caret_predictions <- predict(enet_caret, newdata = X_test)

# Calculate test RMSE
caret_test_rmse <- sqrt(mean((y_test - caret_predictions)^2))

cat("caret Elastic Net Test RMSE: $", 
    format(round(caret_test_rmse), big.mark = ","), "\n", sep = "")

# Compare to manual implementation
cat("Manual Elastic Net Test RMSE: $",
    format(round(enet_test_rmse), big.mark = ","), "\n", sep = "")
```

**Should be similar** to our manual Elastic Net from Part 4

---

# Comparing Multiple Methods

```{r Part5_multiple_methods}
# Train multiple models with caret
methods <- c("lm", "glmnet")  # OLS and Elastic Net
results_list <- list()

for (method in methods) {
  set.seed(123)
  
  if (method == "glmnet") {
    model <- train(X_train, y_train, 
                   method = method,
                   trControl = train_control,
                   tuneGrid = tune_grid)
  } else {
    model <- train(X_train, y_train,
                   method = method,
                   trControl = train_control)
  }
  
  results_list[[method]] <- model
}
```

**Unified interface:** Same code structure for any model!

---

# Model Comparison

```{r Part5_compare_models}
# Compare models
comparison <- resamples(list(
  OLS = results_list$lm,
  Elastic_Net = results_list$glmnet
))

summary(comparison)
```

**resamples()** compares cross-validation results

---

# Visualization of Comparison

```{r Part5_compare_viz, fig.height=4.5}
# Boxplot comparison
bwplot(comparison, 
       metric = "RMSE",
       main = "Cross-Validation RMSE Comparison")
```

**Lower boxes = Better performance**

---

# Extracting Final Model

```{r Part5_final_model}
# Get the underlying glmnet model
final_glmnet <- enet_caret$finalModel

# Extract coefficients
final_coefs <- coef(final_glmnet, 
                    s = best_params$lambda)

print(final_coefs)

# Count non-zero features
n_features <- sum(final_coefs != 0) - 1
cat("\nFeatures selected:", n_features, "\n")
```

**finalModel:** The actual glmnet object with best parameters

---

# Saving the caret Model

```{r Part5_save_caret, eval=FALSE}
# Save complete caret model
saveRDS(enet_caret, "models/enet_caret_model.rds")

# Save just what's needed for production
production_bundle <- list(
  model = enet_caret,
  best_alpha = best_params$alpha,
  best_lambda = best_params$lambda,
  feature_names = colnames(X_train),
  selected_features = rownames(final_coefs)[final_coefs != 0],
  training_performance = enet_caret$results %>%
    filter(alpha == best_params$alpha, 
           lambda == best_params$lambda)
)

saveRDS(production_bundle, "models/production_model.rds")
```

---

# Loading and Using Saved Model

```{r Part5_load_model, eval=FALSE}
# Load the model
loaded_model <- readRDS("models/enet_caret_model.rds")

# Make predictions on new data
new_predictions <- predict(loaded_model, newdata = new_data_matrix)

# Get model info
print(loaded_model$bestTune)
print(loaded_model$results)
```

**Production-ready:** Everything needed is saved

---

# Decision Framework

```{r Part5_decision_framework, echo=FALSE, fig.height=5}
library(DiagrammeR)

grViz("
digraph decision {
  graph [rankdir = TB, fontsize = 12]
  
  node [shape = box, style = filled, fillcolor = lightblue, fontsize = 11]
  start [label = 'Start: Building Regression Model']
  ols [label = 'OLS Regression\n✓ Simple\n✓ Interpretable\n✓ Fast']
  ridge [label = 'Ridge Regression\n✓ All features matter\n✓ Handles correlation\n✓ Stable predictions']
  lasso [label = 'LASSO Regression\n✓ Feature selection\n✓ Interpretable\n✓ Simple model']
  enet [label = 'Elastic Net\n✓ Best of both\n✓ Handles correlation\n✓ Feature selection']
  caret_tune [label = 'Use caret for\nautomatic tuning\n✓ Grid search\n✓ CV built-in']
  
  node [shape = diamond, fillcolor = lightyellow, fontsize = 10]
  q1 [label = 'Many features\nor correlation?']
  q2 [label = 'Need feature\nselection?']
  q3 [label = 'Features\ncorrelated?']
  q4 [label = 'Unsure about\nhyperparameters?']
  
  start -> q1
  q1 -> ols [label = 'No']
  q1 -> q2 [label = 'Yes']
  q2 -> q3 [label = 'Yes']
  q2 -> ridge [label = 'No']
  q3 -> enet [label = 'Yes']
  q3 -> lasso [label = 'No']
  
  ols -> q4 [label = 'Done', style = dashed]
  ridge -> q4 [label = 'Done', style = dashed]
  lasso -> q4 [label = 'Done', style = dashed]
  enet -> q4 [label = 'Done', style = dashed]
  
  q4 -> caret_tune [label = 'Yes']
}
", height = 500)
```

---

# When to Use Each Method

**Decision Table:**

| Situation | Method | Reason |
|-----------|--------|--------|
| Few features, no correlation | OLS | Simple is best |
| Many correlated features | Ridge | Handles correlation |
| Many irrelevant features | LASSO | Feature selection |
| Correlated + need selection | Elastic Net | Best of both |
| Unsure about α or λ | caret + grid search | Automatic tuning |
| Production deployment | caret workflow | Standardized pipeline |

---

# Practical Recommendations

**For RetailCorp Final Model:**

1. ✅ **Use Elastic Net** (balances all needs)
2. ✅ **Use caret for tuning** (finds optimal α and λ)
3. ✅ **Save complete bundle** (model + metadata)
4. ✅ **Monitor in production** (track RMSE monthly)
5. ✅ **Retrain quarterly** (as new data arrives)

**Why this approach?**
- Automated hyperparameter selection
- Reproducible workflow
- Easy to deploy and maintain
- Professional standard

---

# Complete Comparison Table

```{r Part5_complete_table}
# Final comparison of all approaches
final_table <- tibble(
  Method = c("OLS", "Ridge (manual)", "LASSO (manual)", 
             "Elastic Net (manual)", "Elastic Net (caret)"),
  Alpha = c("N/A", "0", "1", "0.5", paste(best_params$alpha)),
  Lambda = c("N/A", "CV selected", "CV selected", "CV selected", 
             paste(round(best_params$lambda, 2))),
  Features = c(10, 10, n_selected, n_selected_enet, n_features),
  Test_RMSE = c(61247, ridge_test_rmse, lasso_test_rmse, 
                enet_test_rmse, caret_test_rmse),
  Tuning = c("None", "Manual CV", "Manual CV", "Manual CV", 
             "Auto grid search")
) %>%
  mutate(Test_RMSE = round(Test_RMSE))

kable(final_table, 
      caption = "Complete Method Comparison",
      format.args = list(big.mark = ","))
```

---

# Performance Visualization

```{r Part5_final_performance, echo=FALSE, fig.height=4.5}
final_table %>%
  ggplot(aes(x = reorder(Method, -Test_RMSE), y = Test_RMSE, fill = Method)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = paste0("$", format(Test_RMSE, big.mark = ","))),
            vjust = -0.5, size = 4) +
  coord_flip() +
  labs(title = "Final Model Comparison: Test RMSE",
       subtitle = "Lower is better - All regularized methods perform similarly",
       x = NULL, y = "Test RMSE") +
  scale_y_continuous(labels = scales::dollar) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

---

# Key Takeaways: caret Integration

**Benefits of caret:**
1. ✅ Unified interface for 238+ models
2. ✅ Automatic hyperparameter tuning
3. ✅ Built-in cross-validation
4. ✅ Easy model comparison
5. ✅ Standardized workflow
6. ✅ Variable importance extraction
7. ✅ Production-ready pipelines

**Cost:** Slightly longer training time (but worth it!)

---

# Production Deployment Checklist

**Before deploying to production:**

- [ ] Model trained with cross-validation
- [ ] Hyperparameters optimized
- [ ] Test RMSE acceptable to stakeholders
- [ ] Feature stability verified
- [ ] Model saved with metadata
- [ ] Documentation complete
- [ ] Monitoring plan in place
- [ ] Retraining schedule defined
- [ ] Stakeholder sign-off obtained

---

# Monitoring in Production

```{r Part5_monitoring, eval=FALSE}
# Monthly monitoring script
monitor_model <- function(new_data, model, threshold = 60000) {
  # Make predictions
  predictions <- predict(model, newdata = new_data)
  
  # Calculate RMSE
  actual_rmse <- sqrt(mean((new_data$actual - predictions)^2))
  
  # Check performance
  if (actual_rmse > threshold) {
    warning("Model performance degraded! RMSE: ", actual_rmse)
    # Trigger retraining
  }
  
  # Log results
  log_entry <- tibble(
    date = Sys.Date(),
    rmse = actual_rmse,
    n_predictions = nrow(new_data),
    alert = actual_rmse > threshold
  )
  
  write_csv(log_entry, "logs/model_performance.csv", append = TRUE)
}
```

---

# Model Retraining Strategy

**When to retrain:**

1. **Scheduled:** Every 3-6 months
2. **Performance-triggered:** RMSE > threshold
3. **Data drift:** Feature distributions change
4. **Business changes:** New store types, markets
5. **After major events:** Economic shifts, pandemics

**Retraining process:**
- Combine old + new data
- Re-run caret tuning
- Compare to previous model
- A/B test before full deployment
- Document changes

---

# Advanced: Nested Cross-Validation

```{r Part5_nested_cv, eval=FALSE}
# Nested CV for unbiased performance estimate
# Outer CV: Estimate performance
# Inner CV: Tune hyperparameters

outer_folds <- createFolds(y_train, k = 5)
nested_results <- map_dfr(outer_folds, function(fold) {
  # Split data
  train_idx <- setdiff(1:length(y_train), fold)
  val_idx <- fold
  
  # Inner CV for tuning
  inner_model <- train(
    X_train[train_idx, ], y_train[train_idx],
    method = "glmnet",
    trControl = trainControl(method = "cv", number = 5),
    tuneGrid = tune_grid
  )
  
  # Evaluate on validation fold
  val_pred <- predict(inner_model, X_train[val_idx, ])
  rmse <- sqrt(mean((y_train[val_idx] - val_pred)^2))
  
  tibble(fold = which(outer_folds == fold), rmse = rmse)
})

cat("Nested CV RMSE:", mean(nested_results$rmse))
```

---

# Business Communication Template

**"To: CFO, RetailCorp**

**Subject: Sales Forecasting Model - Final Recommendation**

After comprehensive analysis, we recommend **Elastic Net regression** with the following specifications:

**Model Performance:**
- Test RMSE: $[X,XXX] (within $25K target)
- Cross-validated accuracy: [X]%
- [X] key features identified

**Key Sales Drivers:**
1. Square footage (strongest)
2. Neighborhood quality
3. Location accessibility
[... list selected features ...]

**Implementation:**
- Deployed via caret pipeline
- Monthly monitoring automated
- Quarterly retraining scheduled

**Next Steps:**
- [ ] Stakeholder approval
- [ ] Deploy to staging
- [ ] Monitor for 1 month
- [ ] Full production rollout"

---

# Summary: Complete Journey

**What we've learned across 5 Partes:**

**Part 1:** Problem diagnosis
- Overfitting detection
- Multicollinearity via VIF

**Part 2:** Ridge regression
- L2 penalty, handles correlation

**Part 3:** LASSO regression  
- L1 penalty, feature selection

**Part 4:** Elastic Net
- Combined L1 + L2, balanced approach

**Part 5:** Professional workflows
- caret integration, production deployment

---

# Skills You've Mastered

**Technical:**
- ✅ Train/test splitting
- ✅ Cross-validation
- ✅ Hyperparameter tuning
- ✅ Ridge, LASSO, Elastic Net
- ✅ caret workflows
- ✅ Model comparison
- ✅ Feature importance

**Professional:**
- ✅ Problem diagnosis
- ✅ Method selection
- ✅ Business communication
- ✅ Production deployment
- ✅ Model monitoring

**You're now a regularization expert!** 🎓

---

# Resources for Continued Learning

**Books:**
- *Introduction to Statistical Learning* (James et al.)
- *Applied Predictive Modeling* (Kuhn & Johnson)
- *Feature Engineering and Selection* (Kuhn & Johnson)

**Online:**
- caret documentation: http://topepo.github.io/caret/
- glmnet vignette
- DataCamp courses on regularization

**Practice:**
- Kaggle competitions
- UCI Machine Learning Repository
- Your own data projects!

---

# Final Recommendations

**For your projects:**

1. **Start simple** (OLS) then add complexity
2. **Always validate** on holdout data
3. **Use caret** for professional workflows
4. **Document everything** for reproducibility
5. **Monitor in production** for model drift
6. **Keep learning** new techniques

**Remember:** 
- Simple is often better
- Understand before implementing
- Communicate clearly with stakeholders
- Models are tools, not magic

---

class: inverse, center, middle

# Congratulations! 🎉

## You've Completed Advanced Regression & Regularization

**You're ready to build production-grade models!**

---

# Questions?

**Topics covered:**
- Overfitting and multicollinearity
- Ridge, LASSO, Elastic Net
- caret workflows
- Production deployment

**Office Hours:** [Your time/location]

**Next Lecture:** Tree-Based Methods & Ensemble Learning

---

class: center, middle

# Thank You! 🙏

**Keep practicing and exploring!**

---

