---
title: "Classwork 5: Advanced Topics - caret Integration"
subtitle: "Professional ML Workflows"
author: "Your Name Here"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

library(tidyverse)
library(glmnet)
library(caret)
library(knitr)

theme_set(theme_minimal(base_size = 14))
```

---

# Introduction

**Scenario:** You're now a senior data scientist at **HomeSmart Realty**. Your models (Ridge, LASSO, Elastic Net) work well, but your manager wants:

> "Can you create a production-ready pipeline that:
> 1. Automatically tunes hyperparameters
> 2. Compares multiple models systematically
> 3. Is easy to deploy and maintain"

**Your task:** Use the `caret` package to create professional ML workflows.

---

# What is caret?

**caret** = Classification And REgression Training

**Key benefits:**
- **Unified interface** for 238+ models
- **Automatic tuning** via grid search
- **Built-in cross-validation**
- **Easy model comparison**
- **Production-ready** workflows

**Philosophy:** Learn one syntax, use any model!

---

# Load the Data

```{r load_data}
# Use the housing data from previous classworks
set.seed(42)
n <- 500

housing_data <- tibble(
  house_id = 1:n,
  square_feet = rnorm(n, 2000, 500),
  bedrooms = sample(2:6, n, replace = TRUE),
  bathrooms = sample(1:4, n, replace = TRUE),
  lot_size = rnorm(n, 8000, 2000),
  age_years = sample(1:50, n, replace = TRUE),
  garage_spaces = sample(0:3, n, replace = TRUE),
  school_rating = rnorm(n, 7, 1.5),
  crime_index = rnorm(n, 50, 15),
  distance_downtown = rnorm(n, 5, 2),
  nearby_parks = sample(0:5, n, replace = TRUE),
  hoa_fees = rnorm(n, 200, 100),
  property_tax = rnorm(n, 3000, 800)
) %>%
  mutate(
    price = 100000 + 
      150 * square_feet + 
      20000 * bedrooms +
      15000 * bathrooms +
      10 * lot_size +
      -2000 * age_years +
      10000 * garage_spaces +
      8000 * school_rating +
      -500 * crime_index +
      -3000 * distance_downtown +
      5000 * nearby_parks +
      rnorm(n, 0, 30000)
  ) %>%
  mutate(price = pmax(price, 50000))

# Prepare matrices
set.seed(123)
train_idx <- createDataPartition(housing_data$price, p = 0.8, list = FALSE)

X_train <- model.matrix(price ~ . - house_id, data = housing_data[train_idx, ])[, -1]
y_train <- housing_data$price[train_idx]

X_test <- model.matrix(price ~ . - house_id, data = housing_data[-train_idx, ])[, -1]
y_test <- housing_data$price[-train_idx]

cat("Data ready!\n")
```

---

# Exercise 1: Set Up Training Control

**Task:** Define how caret should train and validate models.

```{r ex1}
# YOUR CODE HERE
# Create trainControl object
# Hints:
# - Use method = "cv" for cross-validation
# - Use number = 10 for 10-fold CV
# - Set search = "grid" for grid search




# Print to verify
```

**Question:** What does trainControl specify?

---

# Exercise 2: Define Tuning Grid

**Task:** Create a grid of alpha and lambda values for Elastic Net.

```{r ex2}
# YOUR CODE HERE
# Create tuning grid
# Hints:
# - Use expand.grid()
# - Alpha: 0, 0.2, 0.4, 0.6, 0.8, 1.0
# - Lambda: 10^seq(5, -2, length = 50)




# View first few rows


# How many combinations total?
```

**Question:** How many alpha-lambda combinations will caret test?

---

# Exercise 3: Train Elastic Net with caret

**Task:** Use caret's train() function to fit Elastic Net and find optimal hyperparameters.

```{r ex3}
# YOUR CODE HERE
set.seed(123)
# Use train() function
# Hints:
# - method = "glmnet"
# - trControl = your training control
# - tuneGrid = your tuning grid
# - metric = "RMSE"

# This may take 1-2 minutes...




# Print results


# What are the best hyperparameters?
```

**Question:** What alpha and lambda did caret select as optimal?

---

# Exercise 4: Visualize Tuning Results

**Task:** Plot how performance varies across the parameter space.

```{r ex4}
# YOUR CODE HERE
# Use plot() on your caret model




# Try different plot types:
# plot(model, plotType = "level")  # Heatmap
# plot(model, plotType = "scatter") # Line plot
```

**Question:** What patterns do you see? Which alpha values work best?

---

# Exercise 5: Extract Variable Importance

**Task:** Use caret's varImp() to identify the most important features.

```{r ex5}
# YOUR CODE HERE
# Get variable importance




# Print importance values


# Plot importance


# Top 5 features?
```

**Question:** What are the top 5 most important features?

---

# Exercise 6: Compare Multiple Models

**Task:** Train and compare OLS, Ridge, LASSO, and Elastic Net using caret.

```{r ex6}
# YOUR CODE HERE
# Train multiple models and compare

# 1. Create a list to store models


# 2. Train OLS (method = "lm")


# 3. Train Ridge (method = "glmnet", alpha = 0)


# 4. Train LASSO (method = "glmnet", alpha = 1)


# 5. Train Elastic Net (method = "glmnet", tune alpha & lambda)


# 6. Use resamples() to compare


# 7. Print summary


# 8. Create comparison plots
```

**Question:** Which model performs best according to cross-validation?

---

# Bonus Challenge 1: Extract Final Model

**Task:** Extract the underlying glmnet model from caret and examine coefficients.

```{r bonus1, eval=FALSE}
# YOUR CODE HERE
# Get the final model




# Extract coefficients


# Compare to manual glmnet
```

**Question:** Are the coefficients the same as manual glmnet?

---

# Bonus Challenge 2: Production Deployment Bundle

**Task:** Create a complete bundle for production deployment.

```{r bonus2, eval=FALSE}
# YOUR CODE HERE
# Create production bundle with:
# - Trained model
# - Best hyperparameters  
# - Feature names
# - Performance metrics
# - Training date




# Save to file


# Test loading and predicting
```

**Question:** What information should you save for production?

---

# Bonus Challenge 3: Automated Model Selection

**Task:** Create a function that automatically selects the best model.

```{r bonus3, eval=FALSE}
# YOUR CODE HERE
# Write a function that:
# 1. Trains multiple models
# 2. Compares their performance
# 3. Returns the best model
# 4. Saves everything to a folder

auto_model_selection <- function(X_train, y_train, X_test, y_test) {
  # Your code here
}

# Test the function
```

**Question:** How would you automate model selection in production?

---

# Summary Questions

**Answer these conceptual questions:**

1. **What are the main benefits of using caret?**
   - Your answer:

2. **How does caret's train() function differ from glmnet()?**
   - Your answer:

3. **When would you use caret vs manual glmnet?**
   - Your answer:

4. **What is trainControl and why is it important?**
   - Your answer:

5. **How do you compare multiple models in caret?**
   - Your answer:

---

# Self-Check: Did You...

- [ ] Set up trainControl for cross-validation
- [ ] Create tuning grid for alpha and lambda
- [ ] Train Elastic Net with caret
- [ ] Visualize tuning results
- [ ] Extract variable importance
- [ ] Compare multiple models systematically
- [ ] Answer all questions

---

# Submission Instructions

**To submit your work:**

1. Complete all 6 core exercises
2. Answer all questions
3. Attempt at least 1 bonus challenge
4. Run all code chunks to verify they work
5. Knit this document to HTML
6. Submit the HTML file to Canvas

**Grading Criteria:**
- Code runs without errors (40%)
- Correct implementation (40%)
- Quality of insights (20%)

**Total Points:** 100 (+ bonus)

---

# Expected Results

**Your caret model should:**
- Find optimal alpha around 0.4-0.7
- Find optimal lambda via cross-validation
- Achieve test RMSE around $30,000-35,000
- Identify square_feet as most important feature
- Perform similarly to or better than manual methods

**Common issues:**
- Training takes longer than manual glmnet (this is normal!)
- Different random seeds give different results
- Large tuning grids can be slow (reduce size if needed)

---

# Key Takeaways

**caret advantages:**
1. ✅ Consistent syntax across 238+ models
2. ✅ Automated hyperparameter tuning
3. ✅ Built-in cross-validation
4. ✅ Easy model comparison
5. ✅ Variable importance extraction
6. ✅ Production-ready workflows

**When to use caret:**
- Building production pipelines
- Comparing many models
- Need automation
- Want standardization

**When to use manual glmnet:**
- Need fine control
- Speed is critical
- Simple one-off analysis

---

**Good luck! 🚀**

**Remember:** caret is used by data science teams worldwide for production ML!
